### 2015-07-05 paper.tex
Add ISIT paper.

\documentclass[conference]{IEEEtran}

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

\setlength{\parskip}{0.05cm}

% *** MATH PACKAGES ***
%
\usepackage[cmex10]{amsmath}
\usepackage{amssymb}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

\usepackage{color}

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\newcommand\numberthis{\stepcounter{equation}\tag{\theequation}}
\newcommand{\m}[1]{\mathbf{#1}^m}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Some thought experiments on the applicability of Granger causality and Directed Information in inferring direction of information flows}
\author{Praveen Venkatesh and Pulkit Grover\\Carnegie Mellon University}

% make the title area
\maketitle

\begin{abstract}
%\boldmath
Not without controversy, Granger causality and, more recently, directed information, are two measures used to estimate ``causal influence'' of one stochastic process on another. Recently, many works interpret direction of causal influence as the direction of ``information flow.'' To test this latter interpretation, this paper uses two simple experiments to test whether these measures predict the correct direction of information flow, \textit{i.e.}, from the transmitting to the receiving node. These measures have been used extensively in neuroscience, and complex interconnected systems in general, to infer causal influences, and more recently, to infer direction of information flows. Using the problem of communication across feedback channels (first with noiseless feedback, and then with noisy feedback), and using a strategy inspired by the work of Schalkwijk and Kailath, we show that the direction of information flow can be opposite to that inferred using Granger causality and Directed Information. Thus, while it might be reasonable to infer direction of causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow. 
\end{abstract}
\IEEEpeerreviewmaketitle

\section{Introduction}
\label{sec:intro}

This work is in large part motivated by a recent surge of interest in understanding neural circuits --- the connectivity and dynamic activity of parts of the brain --- and how they give rise to behavior and experience\footnote{This is evidenced in the launching of the BRAIN initiative in the US and the Human Brain Project in Europe.}. The question of how the brain computes has been of immense interest for several decades, but only recently have measurement techniques become sophisticated enough to be able to simultaneously record the activity of multiple neurons, or multiple neural populations.

In order to understand how the brain performs computations, it could be useful to understand first the direction of information flows  in various parts of the brain (e.g.~\cite{blinowska2004granger,dhamala2008analyzing,nolte2008robustly,korzeniewska2003determination,schippers2010mapping} etc.). For instance, what one might seek is a component-wise block diagram of the brain, where the brain is broken down into discrete computational units. In this picture, each computational unit receives information, processes it, and then passes it on to another unit.

In order to arrive at such an understanding of the brain, a natural method is to probe it to understand its functional connectivity\footnote{This is different from structural or anatomical connectivity in that links may sometimes be present, but not active.}. Analyses are then performed on the time series thus obtained from the probes, in order to say something about the functional connectivity --- which components influence each other, and which direction is this influence in. One commonly used analysis is based on the computation of Granger causality~\cite{Granger} with the measurements from each node. The Granger causality between two processes measures the level of influence that one process $\{X\}$ has on another process $\{Y\}$. The analysis compares the \textit{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$. Many variations of Granger causality have also been considered, including a generalization --- the Directed Information (see~\cite{MasseyCausality,quinn,weissman}) --- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$. %So another strategy that could be used to infer causal influence of one process, e.g. $\{X\}$, on another process $\{Y\}$ is to estimate and compare $I(\m{X}\to\m{Y})$ and $I(\m{Y}\to\m{X})$ (normalized by $m$) in the limit of large $m$.

The question of when Granger causality is true causality has been hotly debated for quite some time (see, e.g.~\cite{jacobs1979difficulties} for some of the issues). Nevertheless, Granger causality is widely used for inferring causal influence in a wide variety of applications, ranging from biology to economics. In this work, we are more interested in a different issue on use of Granger causality (and Directed Information) that has received little attention: the interpretation of the dominant direction of causal influence as the direction of information flow. Quoting~\cite{nolte2008robustly} for instance, ``To understand interacting systems it is of fundamental importance to distinguish the \emph{driver from the recipient}, and hence to be able to estimate the direction of information flow. The direction can be estimated with a temporal argument: the driver is earlier than the recipient implying that the driver contains information about the future of the recipient not contained in the past of the recipient while the reverse is not the case'' (emphasis has been added by us). Is the dominant direction of directed information really the direction of information flow? Can we assign communication labels of ``transmitter'' and ``receiver'' along the direction of directed information?

%Granger causality and Directed Information can, then, be used to estimate the direction of ``information flow'', and Granger causality often is. But in order to achieve an understanding of how the brain computes, it is necessary to think of the brain regions as communicating, as from a transmitter to a receiver. The authors of this paper believe that ascribing a direction to information flow in the brain, towards a goal of understanding how it computes, is meaningful only if there is also an implicit ``labelling'' of the brain regions being discussed as ``Trasmitter'' and ``Receiver''.

%In this paper, we question whether one is justified in assigning labels of ``Transmitter'' and ``Receiver'', based only on the activity we record from probes? 
%As one might anticipate, such an assignment can be a somewhat ambiguous for a system with feedback loops. 
This paper considers two experiments (introduced in Section~\ref{sec:sk}) where a transmitter $Tx$ wants to communicate a message to a receiver $Rx$ in presence of a feedback channel (in one experiment, the feedback link is noiseless, while in the other it is noisy). We assume that the experimenter is able to record the transmissions of $Tx$ and $Rx$ using some probing mechanism. Provided with these measurements, the experimenter wants to estimate which node is $Tx$ and which is $Rx$. Our results, derived in Section~\ref{sec:analytical-results} and numerically illustrated in Section~\ref{sec:numerical-results}, show that both Granger causality and Directed Information can incorrectly estimate which node is which. The first experiment is for the (unrealistic) case of communicating across noiseless feedback channels. The second experiment allows for noise in the feedback channel as well. In both cases, linear strategies inspired from the strategy of Schalkwijk and Kailath~\cite{S&K} are used.

Our goal here is to bring out the point that whether Granger causality and Directed Information can be used to interpret direction of information flows is an issue that can be, and perhaps should be, considered using thought experiments on simple communication problems where \textit{information flow direction, and quantity, is already known}. If the prediction of Granger causality, or some other measure of information flow, matches with the known information flow, then that measure can be more confidently used in experiments. While our results strongly suggest that one needs to exercise care in interpreting Granger causality and Directed Information dominance as an indicator for the direction of information flow, there are several shortcomings that need to be addressed in order to understand the issue at depth. These shortcomings are discussed in detail in Section~\ref{sec:conclusions}.

%\begin{itemize}

%\item Motivation: Recent interest in understanding neural circuits, to understand how computation is being done -- the BRAIN initiative in the US and the Blue Brain (?) project for mapping the brain in Europe. The ultimate goal is to understand how the brain works, and potentially to use that to engineer better circuits and systems (if we can quote from either of the projects, even better!). Another motivation for this work is probing a circuit to understand how computation is being performed. Or probing a social network?

%\item Leads us to the tale of: What if a biologist came across a radio. How would they examine it? How will they probe it and based on the data, what conclusions will they draw? This has a circuit parallel as well.

%\item The question of causality has been hotly debated elsewhere (see, e.g. ??). More recently, causality has come to be interpreted as the direction of information flows. We need as many citations for this as possible. Some of them: Korzeniewska2003 and papers that cite it. The more papers we can cite with info-flow directions in their title, the better.

%\item In communication parlance, information flows from the transmitter to the receiver. Feedback from the receiver can be viewed as information-flow as well. However, this flow could be of little intellectual interest [... why?...] %information flows not only from transmitter to receiver, but also on the feedback link in the reverse direction. Biological links, as well as circuits, tend to have substantial feedback, and thus it is of great interest to develop techniques to assign the ``transmitting'' and ``receiving'' nodes.

%\item This paper performs some experiments on inferring the direction of information flows. To be able to ask the question of whether the information flow is predicted in the correct direction, we

%\item The first experiment is for the case with noiseless feedback. In this highly asymmetric case, we observe that directed information and Granger causality can incorrectly predict the direction of forward and reverse links.

%\end{itemize}

%Possible shortcomings of our approach/objections to our approach:
%\begin{itemize}
%\item The source is not an ergodic process. However, Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process. So we compute the obvious generalization of Granger causality to non-ergodic processes. 
%\item In feedback communication channels, info flows not just from the transmitter to the receiver, but also from the receiver to the transmitter. So as long as the biologist does not make the leap from direction of information flow to assigning labels such as ``transmitter,'' and ``receiver,'' they will be fine. But if they do not assign these labels, how is computing direction of info flow helping them understand brain circuits? This is unclear to us. It could help them, for instance, to understand epileptic seizure propagation. But that provides little insight into computation (besides, that is really not ``info-flow'' at all!). 
%\item Directed information is computed for a Gaussian case. 
%\item Neural models for spikes tend to be very different from those used here. 
%\item Power and energy constraints are simplified.
%\item We consider a simple point-to-point network. In general networks, this issue could be even more complex.
%\item Granger's original analysis does not compare Granger causality on forward and reverse links. We are doing so, in part because it is often empirically done. Also, the ``conservation of directed information,'' an equality first derived by Massey~\cite{??}, could be viewed as suggesting that an increase in directed info in one direction must reduce it in the other direction. This is not true. For instance, making $X$ and $Y$ uncorrelated reduces directed info in both directions (and their sum) to zero. In other words, Massey's law is not a conservation law in the sense of conservations laws of energy, momentum, or mass, where the total energy/momentum/mass (\textit{i.e.}, the RHS) remains conserved (\textit{i.e.}, constant) regardless of the experiment performed on the entities. 
%\item More counterexamples need to be studied to get a better sense of when these tools are applicable and when they're not.
%\item The work that looks at noisy versions of observed transmissions already creates a concern on use of info flows. Our work bridges that work with info theory. Instead of being a one-off result, that work points to a deeper issue with use of directed info in interpreting info flows. 
%\end{itemize}

%Notes:

%You only seem to be assuming that the variance is $\sigma_\theta^2$. Nothing about whether it is a uniform random variable or something derived from quantizing a uniform random variable, or a Gaussian one.

% --------------------------------------------------------------------------- %

\section{A simple feedback communication scheme}
\label{sec:sk}
This section summarizes the analysis in the work of Schalkwijk and Kailath~\cite{S&K}, and a simple (and previously known) extension to a noisy feedback case. It also establishes the model and the notation used in the paper.

The transmitter wants to convey a single zero-mean random number, $\Theta$, having variance
$\sigma_\theta^2$, to the receiver. $\Theta$ could be obtained, for instance, by quantizing a bounded interval on the real line (e.g. $[-1,1]$), as is done in the scheme of Schalkwijk and Kailath~\cite{S&K}. The forward channel is an AWGN channel with noise variance $\sigma_N^2$. We will use simple linear communication strategies for a noiseless feedback channel, as well as an AWGN feedback channel with noise variance $\sigma_R^2$. In both cases, the estimators will be shown to be unbiased and consistent (the error mean is zero, and the error variance converges to zero).

\subsection{Noiseless feedback: the Schalkwijk-Kailath strategy}
 In the first step, the transmitter sends\footnote{We assume that the power constraints are such that the scaling constant $\alpha$ in~\cite{S&K} is $1$.} $X_1 = \Theta$, which the receiver receives with added noise. The receiver sends back an estimate of $\Theta$ over the feedback link. In all subsequent iterations, the transmitter sends the receiver the error in its latest estimate.

Therefore, in general, the transmitter sends $X_i = \Theta - \widehat\Theta_{i-1}$ and the receiver receives $Y_i = X_i + N_i$, where $N_i \sim \mathcal{N}(0, \sigma_N^2)$ iid. The receiver then estimates $\widehat\Theta_i = \widehat\Theta_{i-1} + \frac{Y_i}{i}$, which results in:
\begin{align*}
	\widehat{\Theta}_{i}  &= \widehat{\Theta}_{i-1}+\frac{X_{i}+N_{i}}{i} \\
					  &= \widehat{\Theta}_{i-1}+\frac{\Theta-\widehat{\Theta}_{i-1}+N_{i}}{i} \\
					  &= \frac{(i-1)\widehat{\Theta}_{i-1}+\Theta+N_{i}}{i} \\
	i\widehat{\Theta}_{i} &= (i-1)\widehat{\Theta}_{i-1}+\Theta+N_{i} \\
					  &=(i-2)\widehat{\Theta}_{i-2}+\Theta+N_{i-1}+\Theta+N_{i} \\
					  & \qquad \vdots \\
					  &= i\Theta+\sum_{j=1}^{i}N_{j} \\
	\widehat{\Theta}_{i}  &=\Theta+\frac{1}{i}\sum_{j=1}^{i}N_{j} \numberthis \label{eq:theta_hat_i-noiseless}
\end{align*}

Through this scheme, the estimate $\widehat{\Theta}_{i}$ is seen to converge to $\Theta$ in mean-square sense as $i\rightarrow\infty$:
\begin{align*}
	\mathbb{E}[\widehat{\Theta}_{i}] &= \mathbb{E}\bigg[\Theta+\frac{1}{i}\sum_{j=1}^{i}N_{j}\bigg] = \mathbb{E}[\Theta]+0 \\
	\mathbb{E}[(\widehat{\Theta}_{i}-\Theta)^{2}] &= \mathbb{E}\bigg[\bigg(\frac{1}{i}\sum_{j=1}^{i}N_{j}\bigg)^{2}\bigg] \\
												  &= \frac{1}{i^{2}}\mathbb{E}\bigg[\bigg(\sum_{j=1}^{i}N_{j}\bigg)^{2}\bigg] = \frac{i}{i^{2}}\mathbb{E}[N_{1}^{2}] = \frac{\sigma_{N}^{2}}{i}\overset{i\to\infty}\longrightarrow 0.
\end{align*}

\subsection{Noisy feedback}
In the presence of noise in the feedback link, restricting our attention to linear strategies, we can use a simple modification of the Schalkwijk-Kailath strategy, incorporating the feedback\footnote{We note that implicitly, this strategy assumes an energy/SNR constraint on the feedback link. This is because the receiver simply sends back the estimate, which is shown to converge to the true value of $\Theta$.}. The receiver still simply transmits the estimate $\widehat{\Theta}_i$ based on the $i$-th forward channel output $Y_i=X_i+N_i$. The transmitter now receives corrupted versions $Z_i=\widehat\Theta_{i-1} + R_{i-1}$ of the receiver's transmissions. That is,
\begin{align*}
	\text{Transmitter's transmissions:}\;X_i &= \Theta - (\widehat\Theta_{i-1} + R_{i-1}) \\
	\text{Channel outputs at the receiver:}\;	Y_i &= X_i + N_i \\
	\text{Receiver's estimates \& transmissions:}\;	\widehat\Theta_i &= \widehat\Theta_{i-1} + \frac{Y_i}{i}
\end{align*}
where $R_{i-1}$ is the AWGN noise in the reverse link. $R_i \sim \mathcal{N}(0, \sigma_R^2)$ are iid random variables.

Linear strategies are known to be suboptimal for the communication problem~\cite{YoungHanKimPaper} (where $\Theta$ is a quantized random variable communicating a finite-rate message reliably), and for problems with non-classical information structures in general~\cite{Witsenhausen68}. Nevertheless, we now show that the resulting estimates $\widehat{\Theta}_i$  still converge to $\Theta$ in mean-square sense:
\begin{align*}
	i\widehat\Theta_i &= i\widehat\Theta_{i-1} + X_i + N_i \\
					  &= i\widehat\Theta_{i-1} + \Theta - \widehat\Theta_{i-1} - R_{i-1} + N_i \\
					  &= (i-1) \widehat\Theta_{i-1} + \Theta - R_{i-1} + N_i \numberthis \label{eq:theta_hat_i-noisy-once} \\
					  &\overset{(a)}{=} i\Theta + \sum_{k=1}^i N_k - \sum_{k=1}^{i-1} R_k \\
	\widehat\Theta_i  &= \Theta + \frac{1}{i} \sum_{k=1}^i N_k - \frac{1}{i} \sum_{k=1}^{i-1} R_k \numberthis \label{eq:theta_hat_i-noisy}
\end{align*}
where $(a)$ is obtained by expanding $\widehat{\Theta}_j$ recursively for $j=i-1,i-2,\ldots,2$. Therefore, the error in estimating $\Theta$ converges to $0$ in mean-square sense (\textit{i.e.}, $\widehat{\Theta}_i$ is a consistent estimate of $\Theta_i$ even in the presence of noise on the feedback link).

We also derive here an expression for $X_i$, which we will use later.
\begin{align*}
	X_i &= \Theta - \widehat\Theta_{i-1} - R_{i-1} \\
	&= \Theta - \bigg( \Theta + \frac{1}{i-1} \sum_{k=1}^{i-1} N_k - \frac{1}{i-1} \sum_{k=1}^{i-2} R_k \bigg) - R_{i-1} \\
	&= \frac{1}{i-1} \sum_{k=1}^{i-2} R_k - \frac{1}{i-1} \sum_{k=1}^{i-1} N_k - R_{i} \numberthis \label{eq:x_i-noisy}
\end{align*}

\section{Granger causality and directed information analyses for the strategies in Section~\ref{sec:sk}}
\label{sec:analytical-results}

\subsection{Granger causality for the noiseless feedback case}

Suppose we try to use the Granger causality measure of directed influence to infer the direction of information flow. As discussed in Section~\ref{sec:intro}, one might expect the direction of information flow --- and hence the direction of greater Granger causality --- to be from the transmitter to the receiver.

In order to compute the Granger causality in the reverse direction (from receiver to transmitter), we wish first to compute coefficients $\alpha_j$ such that:
\begin{equation*}
	X_i = \sum_{j=1}^{p}{\alpha_j X_{i-j}} + \epsilon_i
\end{equation*}
Note that $\alpha_j$ can themselves depend on $i$, since this is a non-stationary process (because the error $\Theta-\widehat{\Theta}_i = X_i$ converges to zero).
\begin{align*}
	X_i &= \Theta - \widehat\Theta_{i-1} \\
		&= \Theta - \big( \widehat\Theta_{i-2} + \frac{Y_{i-1}}{i-1} \big) \\
		&= \Theta - (\Theta - X_{i-1}) - \frac{X_{i-1} + N_{i-1}}{i-1} \\
		&= X_{i-1} - \frac{X_{i-1} + N_{i-1}}{i-1} \\
		&= \frac{i-2}{i-1} X_{i-1} + \frac{N_{i-1}}{i-1}
\end{align*}
Therefore, $\alpha_1 = \frac{i-2}{i-1}$ and $\epsilon_i = \frac{N_{i-1}}{i-1}$, and hence $\text{Var}(\epsilon_i) = \frac{\sigma_N^2}{(i-1)^2}$.

Next, we wish to write $X_i$ in terms of $X_{i-j}$ and $\widehat\Theta_{i-j}$:
\begin{equation*}
	X_i = \sum_{j=1}^{p}{\alpha_j X_{i-j}} + \sum_{j=1}^{p}{\beta_j \widehat\Theta_{i-j}} + \tilde\epsilon_i
\end{equation*}
which gives
\begin{align*}
	X_i &= \Theta - \widehat\Theta_{i-1} \\
		&= X_{i-1} + \widehat\Theta_{i-2} - \widehat\Theta_{i-1}
\end{align*}
Since there is no noise expression here, the Granger causality ratio, $\text{Var}(\epsilon_i) / \text{Var}(\tilde\epsilon_i)$ goes to infinity.

In the forward direction, we do not explicitly compute the Granger causality ratio, but simply show that it is bounded strictly between 0 and $\infty$:
\begin{equation*}
	\widehat\Theta_i = \sum_{j=1}^p{\alpha_j \widehat\Theta_{i-j}} + \epsilon_i
\end{equation*}
\begin{align*}
	\widehat\Theta_i &= \widehat\Theta_{i-1} + \frac{X_i + N_i}{i} \\
					 &= \widehat\Theta_{i-1} + \frac{\Theta - \widehat\Theta_{i-1} + N_i}{i} \\
					 &= \frac{i-1}{i} \widehat\Theta_{i-1} + \frac{\Theta}{i} + \frac{N_i}{i}
\end{align*}
which means that $0 < \frac{\sigma_N^2}{i^2} < \text{Var}(\epsilon_i) < \frac{\sigma_N^2 + \sigma_\theta^2}{i^2} < \infty$.
Further, if we try to predict $\widehat\Theta_i$ from the previous $\widehat\Theta_{i-i}$ and $X_{i-j}$:
\begin{equation*}
	\widehat\Theta_i = \sum_{j=1}^{p}{\alpha_j \widehat\Theta_{i-j}} + \sum_{j=0}^{p-1}{\beta_j X_{i-j}} + \tilde\epsilon_i
\end{equation*}
we see that
\begin{equation*}
	\widehat\Theta_i = \widehat\Theta_{i-1} + \frac{X_i}{i} + \frac{N_i}{i}
\end{equation*}
so that $\text{Var}(\tilde\epsilon_i) = \sigma_N^2/i^2$. The Granger causality ratio, $\text{Var}(\epsilon_i) / \text{Var}(\tilde\epsilon_i)$, in the forward direction is, therefore, finite.

The intuitive argument for why this is happening might go as follows: since the feedback link is noiseless, you can always perfectly predict the transmitted symbol from the $\widehat\Theta$'s and the history of $X$. On the other hand, you can never perfectly predict $\widehat\Theta_i$ from the $X$'s and the history of $\widehat\Theta$.

\subsection{Directed Information for the noiseless feedback case}

Performing the Directed Information analysis for the scheme described above yields the same results. In order to ease the burden of computing Directed Information, we assume that $\Theta$ is normally distributed.

The directed information in the forward direction is computed as:
\begin{equation*}
	I(X^{n} \rightarrow \widehat{\Theta}^{n}) = \frac{1}{2}\log \bigg( 1+\frac{n\sigma_{\theta}^{2}}{\sigma_{N}^{2}} \bigg)
\end{equation*}
where $n$ is the number of iterations of the Schalkwijk-Kailath algorithm. For a proof, refer appendix~\ref{app:appendix3}.

In the reverse direction, the Directed Information is infinite.
\begin{align*}
	I(0*\widehat{\Theta}^{n-1}\rightarrow X^{n}) &= \sum_{i=0}^{n-1}I(X_{i+1};\widehat{\Theta}^{i}|X^{i}) \\
	I(X_{i+1};\widehat{\Theta}^{i}|X^{i})        &= h(X_{i+1}|X^{i}) - h(X_{i+1}|X^{i},\widehat{\Theta}^{i}) \numberthis \label{eq:dir-info-rev-noiseless}
\end{align*}
The first term in the equation above reduces to
\begin{align*}
	h(X_{i+1}|X^{i}) &= h(\Theta-\widehat{\Theta}_{i}|X^{i}) \\
					 &= h \bigg( \Theta- \bigg( \widehat{\Theta}_{i-1}+\frac{X_{i}+N_{i}}{i} \bigg) \bigg| X^{i} \bigg) \\
					 &\overset{(a)}{=} h \bigg( \Theta- \bigg( (\Theta - X_{i}) + \frac{N_{i}}{i} \bigg) \bigg| X^{i} \bigg) \\
					 &= h \bigg( \frac{N_{i}}{i} \bigg| X^{i} \bigg) \\
					 &= h(N_{i}) -\log(i) \\
					 &= \frac{1}{2}\log(2\pi e\sigma_{N}^{2}) -\log(i)
\end{align*}
where for (a), we have dropped $X_i/i$ from the previous step, since it is conditioned over, and then written $\widehat\Theta_{i-1}$ as $(\Theta - X_i)$. On the other hand, the second term in equation \eqref{eq:dir-info-rev-noiseless} becomes
\begin{align*}
	h(X_{i+1} | X^i, \widehat\Theta^i) &= h(\Theta - \widehat\Theta_i | X^i, \widehat\Theta^i) \\
									   &= h(\Theta | X^i, \widehat\Theta^i) \\
									   &\overset{(a)}{=} h(X_i + \widehat\Theta_{i-1} | X^i, \widehat\Theta^i) \\
									   &= h(0 | X^i, \widehat\Theta^i) \\
									   &\overset{(b)}{=} - \infty
\end{align*}
where for (a) we have expressed $\Theta$ in terms of $\widehat\Theta_{i-1}$ and $X_i$ and for (b), we have used the fact that the differential entropy of a constant (or equivalently, a Gaussian with zero variance) is negative infinity. This means that equation \eqref{eq:dir-info-rev-noiseless} becomes
\begin{align*}
	I(X_{i+1};\widehat{\Theta}^{i}|X^{i}) &= \infty \\
	\Rightarrow I(0*\widehat{\Theta}^{n-1}\rightarrow X^{n}) &= \infty
\end{align*}
as expected.

\subsection{Directed Information for the noisy feedback scenario}

Since a noiseless feedback link is not realistic, we proceed to perform the same Directed Information calculations as above for the feedback link with additive white Gaussian noise of variance $\sigma_R^2$. While we could not derive simple closed form expressions for the Directed Information in the forward and reverse links, we were able to evaluate the expressions numerically. These are plotted in section~\ref{sec:numerical-results}.

The Directed Information in the forward direction can be written as
\begin{align*}
	I(X^n \rightarrow \widehat\Theta^n) &= \sum_{i=1}^n{I(\widehat\Theta_i ; X^i | \widehat\Theta^{i-1})} \\
										&= \sum_{i=1}^n{h(\widehat\Theta_i | \widehat\Theta^{i-1}) - h(\widehat\Theta_i | \widehat\Theta^{i-1}, X^i)} \numberthis \label{eq:dir-info-fwd-noisy} \\
										&= \sum_{i=1}^n \bigg( \frac{1}{2} \log(2 \pi e \text{Var}[\Theta - R_{i-1} + N_i | \widehat\Theta^{i-1}]) \\
										& \qquad \qquad - \frac{1}{2}\log(2 \pi e \sigma_N^2) \bigg)
\end{align*}
For a derivation of this, see appendix~\ref{app:appendix1}.

In the reverse direction,
\begin{align*}
	I(0*\widehat\Theta^{n-1} & \rightarrow X^n) = \sum_{i=0}^{n-1}{I(X_{i+1}; \widehat\Theta^i | X^i)} \\
							 &= \sum_{i=0}^{n-1}{h(X_{i+1} | X^i) - h(X_{i+1} | X^i, \widehat\Theta^i)} \numberthis \label{eq:dir-info-rev-noisy} \\
							 &= \frac{1}{2} \log \bigg( 2 \pi e \frac{\sigma_N^2 + \sigma_R^2}{\sigma_R^2} \bigg) \\
							 & \; \; + \sum_{i=2}^{n-1} \bigg( \frac{1}{2} \log \bigg( 2 \pi e \text{Var} \bigg[ R_{i-1} - \frac{N_i}{i} - R_i \bigg| X^i \bigg] \bigg) \\
							 & \qquad \qquad - \frac{1}{2} \log( 2 \pi e \sigma_R^2 ) \bigg)
\end{align*}
For a derivation of this, see appendix~\ref{app:appendix2}.

% --------------------------------------------------------------------------- %

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.

\section{Numerical results}
\label{sec:numerical-results}
In the noiseless feedback case, the directed information on the forward link is finite, while that on the feedback link is infinite. However, the situation is more interesting when there is noise in the feedback link, as is illustrated in Fig.~\ref{fig:numerical} through numerical calculations of expressions in the last section.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
	\centering
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point20} \\
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point35} \\
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point60}
	\caption{Plots for forward and backward directed information computations for ${\sigma_R^2}/{\sigma_N^2}=0.2$ (top), $0.35$ (center) and $0.6$ (bottom). In each plot, curves for directed information in both directions are illustrated for ratios ${\sigma_\theta^2}/{\sigma_N^2} = 0.25, 0.5,$ and $1$. The x-axis is the number of iterations of message-passing between the transmitter and the receiver. For cases when feedback noise variance $\sigma_R^2$ is moderately smaller than feedforward noise variance $\sigma_N^2$, directed information in the reverse link can dominate that in the forward link. With sufficiently many (albeit sometimes large, as illustrated in the top figure) iterations, directed information in forward information starts to dominate. However, the point at which this happens depends on the (often unknown) ratio of noise in these links. }
		     \label{fig:numerical}
\end{figure}

\section{Conclusions and discussions}
\label{sec:conclusions}

%%%%%%%%%%%%%%%%%%
There are several shortcomings of this work that need to be addressed for a more comprehensive understanding of the issue. First, our information source does not evolve with time, and hence our communication process (inspired from the scheme of Schalkwijk and Kailath~\cite{S&K} is non-ergodic. Since Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process, we compute the obvious generalization of Granger causality to non-ergodic processes. Nevertheless, future work will address a situation with a linear dynamical system as the information source. Second, our experiments restrict themselves to Gaussian noise for simplicity, but neural spiking and spike-rate models for spikes tend to be very different from those used here. This is a clear direction for future work. Third, the power and energy constraints are somewhat oversimplified to make the analysis simpler. This is for simplicity of exposition. A more general analysis is a trivial extension. Finally, we consider a simple point-to-point network. In general networks, this issue could be even more complex. However, since this issue shows up even for the simplest network, we feel that the problem will only be exacerbated when the network is large.

We note that Granger's original analysis does not compare Granger causality on forward and reverse links. We are doing so, in part because it is often done in practice. It strikes us that this is being done because of the ``conservation of directed information'', an equality first derived by Massey~\cite{massey2005conservation}. This equality, which states that the sum of forward and reverse Directed Informations is the mutual information, could be viewed as suggesting that an increase in Directed Information in one direction must reduce it in the other direction. However, this is not true. For e.g., making $X$ and $Y$ uncorrelated reduces Directed Information in both directions (along with their sum) to zero. In other words, Massey's law is not a conservation law in the sense of conservation laws of energy, momentum, or mass, where the total energy/momentum/mass (\textit{i.e.}, the RHS) remains conserved (\textit{i.e.}, constant) regardless of the experiment performed on the entities.

\section*{Acknowledgments}

The authors would like to thank Nicola Elia for discussions that suggested this possible misunderstanding in use of Granger causality and Directed Information in computational systems. Praveen Venkatesh is partially supported by a dean's fellowship. %Pulkit Grover is supported by [...].

\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,MyMainBibliographyISIT15}

\newpage

\appendices

\section{Directed Information in the forward direction, noiseless feedback}
\label{app:appendix3}

The Directed Information in the forward direction is computed as:
\begin{align*}
	I(X^n \rightarrow \widehat\Theta^n) &= \sum_{i=1}^n{I(\widehat\Theta_i ; X^i | \widehat\Theta^{i-1})} \\
										&= \sum_{i=1}^n{h(\widehat\Theta_i | \widehat\Theta^{i-1}) - h(\widehat\Theta_i | \widehat\Theta^{i-1}, X^i)} \numberthis \label{eq:dir-info-fwd-noiseless}
\end{align*}
Taking the first term in \eqref{eq:dir-info-fwd-noiseless},
\begin{align*}
	h(\widehat\Theta_i | \widehat\Theta^{i-1}) &= h \bigg( \widehat\Theta_{i-1} + \frac{X_i + N_i}{i} \bigg| \widehat\Theta^{i-1} \bigg) \\
											   &= h(\Theta - \widehat\Theta_{i-1} + N_i | \widehat\Theta^{i-1}) -\log(i) \\
											   &= h(\Theta + N_i | \widehat\Theta^{i-1}) -\log(i) \\
											   &= h(\Theta + N_i | \widehat\Theta_{i-1}) -\log(i)
\end{align*}
where we have dropped the conditioning on all except $\widehat\Theta_{i-1}$ in the last step. %TODO: Justify. Markov proof is too long to write here.
Define $U = \Theta + N_i$ and $V = \widehat\Theta_{i-1}$. Since all variables are Gaussian, it suffices to find the variance of the conditional distribution $U|V$.
\begin{align*}
	\mathbb{E}[U] = 0,
	\mathbb{E}[V] &= 0, \text{Var}[U] = \sigma_\theta^2 + \sigma_N^2,
	\text{Var}[V] = \sigma_\theta^2+ \frac{\sigma_N^2}{i-1} \\
	\text{Cov}[U,V] &= \mathbb{E}[UV]-\mathbb{E}[U]\mathbb{E}[V] \\
					&= \sigma_\theta^2 \\
	\rho^{2}        &= \frac{\sigma_{\theta}^{4}}{(\sigma_{\theta}^{2}+\sigma_{N}^{2})(\sigma_{\theta}^{2}+\frac{\sigma_{N}^{2}}{(i-1)})}
\end{align*}
\begin{equation*}
	U|V=v\ \sim\ \mathcal{N}\left(\sqrt{\frac{\sigma_{\theta}^{2}+\sigma_{N}^{2}}{\sigma_{\theta}^{2}+\frac{\sigma_{N}^{2}}{i-1}}}\rho v,(1-\rho^{2})(\sigma_{\theta}^{2}+\sigma_{N}^{2})\right)
\end{equation*}
Hence, the entropy of the conditional distribution is
\begin{equation*}
	h(U|V=v)=\frac{1}{2}\log(2\pi e(1-\rho^{2})(\sigma_{\theta}^{2}+\sigma_{N}^{2})) \overset{(a)}{=} h(U|V)
\end{equation*}
where (a) follows because the conditional entropy is independent of $v$. Thus,
\begin{equation}
	h(\widehat{\Theta}_{i}|\widehat{\Theta}^{i-1}) = \frac{1}{2}\log\left(2\pi e\frac{\sigma_{N}^{2}(i\sigma_{\theta}^{2}+\sigma_{N}^{2})}{((i-1)\sigma_{\theta}^{2}+\sigma_{N}^{2})}\right)-\log(i) \label{eq:htheta_theta-noiseless}
\end{equation}
The next term in equation \eqref{eq:dir-info-fwd-noiseless} is
\begin{align*}
	h(\widehat{\Theta}_{i}|\widehat{\Theta}^{i-1},X^{i}) &= h\left(\frac{N_{i}}{i}\bigg|\widehat{\Theta}^{i-1},X^{i}\right) \\
												 &= h(N_{i}) -\log(i) \\
												 &= \frac{1}{2}\log(2\pi e\sigma_{N}^{2}) -\log(i) \numberthis \label{eq:htheta_x_theta-noiseless}
\end{align*}
Putting equations \eqref{eq:htheta_theta-noiseless} and \eqref{eq:htheta_x_theta-noiseless} together, we can compute the forward Directed Information:
\begin{align*}
	I(\widehat{\Theta}_{i};X^{i}|\widehat{\Theta}^{i-1}) &= h(\widehat{\Theta}_{i}|\widehat{\Theta}^{i-1}) - h(\widehat{\Theta}_{i}|\widehat{\Theta}^{i-1},X^{i}) \\
	                                             &= \frac{1}{2}\log\left(\frac{i\sigma_{\theta}^{2}+\sigma_{N}^{2}}{(i-1)\sigma_{\theta}^{2}+\sigma_{N}^{2}}\right) \\
	I(X^{n} \rightarrow \widehat{\Theta}^{n})    &= \sum_{i=1}^{n}I(\widehat{\Theta}_{i};X^{i}|\widehat{\Theta}^{i-1}) \\
												 &\overset{(a)}{=} \frac{1}{2}\log\left(\frac{n\sigma_{\theta}^{2}+\sigma_{N}^{2}}{\sigma_{N}^{2}}\right) \\
	                                             &= \frac{1}{2}\log\left(1+\frac{n\sigma_{\theta}^{2}}{\sigma_{N}^{2}}\right)
\end{align*}
where (a) follows through by expanding out the product inside the logarithms and canceling terms. Clearly, this value is finite.

\section{Directed Information in the forward direction, with noisy feedback}
\label{app:appendix1}

\begin{align*}
	I(X^n \rightarrow \widehat\Theta^n) &= \sum_{i=1}^n{I(\widehat\Theta_i ; X^i | \widehat\Theta^{i-1})} \\
									 &= \sum_{i=1}^n{h(\widehat\Theta_i | \widehat\Theta^{i-1}) - h(\widehat\Theta_i | \widehat\Theta^{i-1}, X^i)}
\end{align*}
Taking the first of the two terms in the above expression,
\begin{align*}
	h(\widehat\Theta_i | \widehat\Theta^{i-1}) &\overset{(a)}{=} h(\widehat\Theta_{i-1} \frac{i-1}{i} + \frac{\Theta}{i} - \frac{R_{i-1}}{i} + \frac{N_i}{i} | \widehat\Theta^{i-1}) \\
											   &= h(\Theta - R_{i-1} + N_i | \widehat\Theta^{i-1}) -\log(i)
\end{align*}
where for (a) we have used equation \eqref{eq:theta_hat_i-noisy-once}. The Markov property no longer holds in this case, but we proceed in the same manner. We define $U = \Theta - R_{i-1} + N_i$ and $\underbar{V} = \widehat\Theta^{i-1}$. Recalling equation \eqref{eq:theta_hat_i-noisy}, for $j \in \{1, \ldots i-1\}$, $p \in \{1, \ldots i-1\}$ and $q \in \{1, \ldots i-1\}$, we have
\begin{align*}
	\mathbb{E}[U] &= 0,\;	\mathbb{E}[\widehat\Theta_j] = 0,\;\\	\mathbb{E}[U^2] &= \sigma_\theta^2 + \sigma_R^2 + \sigma_N^2,\;	\mathbb{E}[U \widehat\Theta_j] = \sigma_\theta^2 \\
	\mathbb{E}[\widehat\Theta_p \widehat\Theta_q] &= \mathbb{E} \bigg[ \bigg( \Theta + \frac{1}{p} \sum_{k=1}^p N_k - \frac{1}{p} \sum_{k=1}^{p-1} R_k \bigg)\\
												  & \qquad \qquad \bigg( \Theta + \frac{1}{q} \sum_{k=1}^q N_k - \frac{1}{q} \sum_{k=1}^{q-1} R_k \bigg) \bigg] \\
												  &= \sigma_\theta^2 + \frac{\min\{p, q\}}{pq} \sigma_N^2 + \frac{\min\{p-1, q-1\}}{pq} \sigma_R^2 \\
	\text{Var}[U|\widehat\Theta^{i-1}] &= \mathbb{E}[U^2] - \mathbb{E}[U \underbar{V}] \mathbb{E}[\underbar{V} \underbar{V}^T]^{-1} \mathbb{E}[\underbar{V} U] \\
	h(U | \widehat\Theta^{i-1}) &= \frac{1}{2}\log(2 \pi e \text{Var}[U | \widehat\Theta^{i-1}]) \numberthis \label{eq:h_theta_theta-noisy}
\end{align*}
We can not derive a simple closed form for this expression, but we have computed it numerically for the plots in Section~\ref{sec:numerical-results}. The second term in equation \eqref{eq:dir-info-fwd-noisy} is
\begin{align*}
	h(\widehat\Theta_i | \widehat\Theta^{i-1}, X^i) &= h \bigg( \widehat\Theta_{i-1} + \frac{X_i + N_i}{i} \bigg| \widehat\Theta^{i-1}, X^i \bigg) \\
													&= h(N_i | \widehat\Theta^{i-1}, X^i) -\log(i) \\
													&= \frac{1}{2}\log(2 \pi e \sigma_N^2) -\log(i) \numberthis \label{eq:h_theta_x_theta-noisy}
\end{align*}
From equations \eqref{eq:h_theta_theta-noisy} and \eqref{eq:h_theta_x_theta-noisy}, we compute the forward-directed information as depicted in Section~\ref{sec:numerical-results}, for different values of $\sigma_\theta^2$.

% you can choose not to have a title for an appendix
% if you want by leaving the argument blank
\section{Directed Information in the reverse direction, with noisy feedback}
\label{app:appendix2}

\begin{align*}
	I(0*\widehat\Theta^{n-1} \rightarrow X^n) &= \sum_{i=0}^{n-1}{I(X_{i+1}; \widehat\Theta^i | X^i)} \\
											  &= \sum_{i=0}^{n-1}{h(X_{i+1} | X^i) - h(X_{i+1} | X^i, \widehat\Theta^i)} \numberthis \label{eq:dir-info-rev-noisy}
\end{align*}
Taking the first term inside the summation,
\begin{align*}
	h(X_{i+1} | X^i) &= h(\Theta - \widehat\Theta_i - R_i | X^i) \\
					&\overset{(a)}{=} h \bigg( \bigg( - \widehat\Theta_{i-1} - \frac{X_i + N_i}{i} \bigg) - R_i \bigg| X^i \bigg) \\
					&\overset{(b)}{=} h \bigg( \bigg( X_i - \Theta + R_{i-1} - \frac{N_i}{i} \bigg) - R_i \bigg| X^i \bigg) \\
					&\overset{(c)}{=} h \bigg( R_{i-1} - \frac{N_i}{i} - R_i \bigg| X^i \bigg)
\end{align*}
where in (a) above, we have dropped $\Theta = X_1$, in (b) we have re-expressed $\widehat\Theta_{i-1}$ in terms of $X_i$, $\Theta$ and $R_{i-1}$, and in (c) we have dropped $X_i$ and $\Theta$ again. As before, define $U = R_{i-1} - \frac{N_i}{i} - R_i$, so that
\begin{equation*}
	\mathbb{E}[U] = 0,\; \; \mathbb{E}[X_j] = 0, \; \; \mathbb{E}[U^2] = 2 \sigma_R^2 + \frac{\sigma_N^2}{i^2}
\end{equation*}
For $i \geq 3$ and $j \in \{3, \ldots i\}$, we can use equation \eqref{eq:x_i-noisy} to see that
\begin{align*}
	\mathbb{E}[U X_j] &= \mathbb{E} \bigg[ \bigg( R_{i-1} - \frac{N_i}{i} - R_i \bigg) \\
					  & \qquad             \bigg( \frac{1}{j-1} \sum_{k=1}^{j-2} R_k - \frac{1}{j-1} \sum_{k=1}^{j-1} N_k - R_{j-1} \bigg) \bigg] \\
					  &= - \mathbb{E}[R_{i-1} R_{j-1}] = - \sigma_R^2 \delta_{ij} \\
	\mathbb{E}[U X_1] &= \mathbb{E} \bigg[ \bigg( R_{i-1} - \frac{N_i}{i} - R_i \bigg) \Theta \bigg] = 0 \; \\
	\mathbb{E}[U X_2] &= \mathbb{E} \bigg[ \bigg( R_{i-1} - \frac{N_i}{i} - R_i \bigg) (\Theta - (\Theta + N_i) - R_1) \bigg] = 0 \; \\
	\mathbb{E}[X_p X_q] &= \mathbb{E} \bigg[ \bigg( \frac{1}{p-1} \sum_{k=1}^{p-2} R_k - \frac{1}{p-1} \sum_{k=1}^{p-1} N_k - R_{p-1} \bigg) \\
						& \qquad             \bigg( \frac{1}{q-1} \sum_{k=1}^{q-2} R_k - \frac{1}{q-1} \sum_{k=1}^{q-1} N_k - R_{q-1} \bigg) \bigg] \\
					    &= \frac{\min\{p-2, q-2\}}{(p-1)(q-1)} \sigma_R^2 + \frac{\min\{p-1, q-1\}}{(p-1)(q-1)} \sigma_N^2 \\
						& \qquad + \sigma_R^2 \delta_{pq} + \frac{1}{p-1} \sigma_R^2 \mathbb{I}_{p>q} + \frac{1}{q-1} \sigma_R^2 \mathbb{I}_{q>p} \\
	\mathbb{E}[X_1 X_1] &= \mathbb{E}[\Theta^2] = \sigma_\theta^2 \\
	\mathbb{E}[X_1 X_j] &= \mathbb{E} \bigg[ \Theta \bigg( \frac{1}{j-1} \sum_{k=1}^{j-2} R_k - \frac{1}{j-1} \sum_{k=1}^{j-1} N_k - R_{j-1} \bigg) \bigg] = 0 \\
	\text{Var}[U|X^i] &= \mathbb{E}[U^2] - \mathbb{E}[U X^i] \mathbb{E}[X^i {X^i}^T]^{-1} \mathbb{E}[X^i U] \\
	h(U | X^i) &= \frac{1}{2}\log(2 \pi e \text{Var}[U | X^i]) \numberthis \label{eq:h_x_x-noisy}
\end{align*}
The above argument can be extended to $i=2$ by letting the final index of the summation term be smaller than the starting index, implying that the whole summation term is simply dropped. The special cases of $i=0$ and $i=1$ are handled at the end. The second term from equation \eqref{eq:dir-info-rev-noisy} becomes
\begin{align*}
	h(X_{i+1} | X^i, \widehat\Theta^i) &= h(\Theta - \widehat\Theta_i - R_i | X^i, \widehat\Theta^i) \\
									   &= \frac{1}{2}\log(2 \pi e \sigma_R^2) \numberthis \label{eq:h_x_x_theta-noisy}
\end{align*}
because $X_1 = \Theta$ and because $R_i$ is independent of all the $X^i$ and $\widehat\Theta^i$. For the special cases of $i=0$ and $i=1$, we solve for the value of mutual information explicitly:
\begin{align*}
	i = 0: & \qquad I(X_1; \widehat\Theta^0 | X^0) &=& \; h(X_1) - h(X_1) = 0 \\
	i = 1: & \qquad I(X_2; \widehat\Theta^1 | X^1) &=& \; h(X_2 | X_1) - h(X_2 | X_1, \widehat\Theta_1) \\
		   & \qquad h(X_2 | X_1)                   &=& \; h(\Theta - \widehat\Theta_1 - R_1 | X^1) \\
		   & \qquad                                &=& \; h(-\Theta - N_1 - R_1 | X_1) \\
		   & \qquad                                &=& \; h(-N_1 - R_1) \\
		   & \qquad                                &=& \; \frac{1}{2} \log(2 \pi e (\sigma_N^2 + \sigma_R^2)) \\
		   & \qquad h(X_2 | X_1, \widehat\Theta_1) &=& \; h(-R_1) = \frac{1}{2} \log(2 \pi e \sigma_R^2) \\
		   & \qquad I(X_2; \widehat\Theta^1 | X^1) &=& \; \frac{1}{2} \log \bigg( 2 \pi e \frac{\sigma_N^2 + \sigma_R^2}{\sigma_R^2} \bigg)
\end{align*}
From equations \eqref{eq:h_x_x-noisy} and \eqref{eq:h_x_x_theta-noisy}, along with the two special cases above, we can now compute the reverse-directed information plotted in Section~\ref{sec:numerical-results}.

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{14}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)

% that's all folks
\end{document}

### 2015-07-05 paper.tex
Add main points for Allerton draft.

@@
 \label{sec:intro}
+
+\subsection{Motivation}
+
+\begin{enumerate}
+	\item Increased interest in understanding (a) functional connectivity maps and (b) computational networks of the brain.
+	\item Evidenced by funding - Brain Initiative, Human Brain project.
+	\item Quote one of the missions of the Brain Initiative - esp. for connectivity maps, but also for computation
+	\item Quote papers that try to understand how *information flows* in the brain. (past references from the ISIT paper - look for some more).
+	\item Last line: "The question we raise is - Do these directed connectivity maps, as determined by directional causal influence measures such as Granger causality, correctly identify the directions along which information flows in the brain?"
+\end{enumerate}
+
+\subsection{How Granger Causality is used in Neuroscience today}
+
+\begin{enumerate}
+	\item We take Brovelli et.\ al.\ and Ding et.\ al.\ as the basis / canonical reference, as this appears to be generally accepted by the community (Cite review papers that cite these two papers; cite Anil Seth's MVGC package paper).
+	\item Explain Granger causality
+	\item Explain the exact method used - including trials, pre-processing steps, etc.\ - as described in the Bressler paper.
+	\item Granger's original analysis does not compare GC on forward and reverse links, however, this appears to be the commonly done in practice (cite Brovelli and review papers if possible)
+	\item State our problem statement again - we take issue not with any of these steps, but with the interpretation that if greater Granger causal influences are seen to be from X to Y, then information must be flowing from X to Y in the brain.
+\end{enumerate}
+
+\subsection{Our counter-examples and main result}
+
+\begin{enumerate}
+	\item Define Granger causal influence and directed info in English again.
+Define information flow (use examples, or define formally?
+Formal definition should come in the math sections).
+Describe the problem again - is the direction of causal influence the same as the direction of information flow?
+	\item Describe the counter-examples in English.
+		\begin{enumerate}
+			\item Transmitter has the message (msg is known at the transmitter before it is known at the receiver - this is what distinguishes Tx from Rx).
+Therefore, the true direction in which information is flowing is *known* to us, by design of the experiment.
+This is the *ground truth*, against which we can compare the results of Granger causality.
+			\item Feedback link is present - we consider two cases: feedback link noiseless, and feedback link noisy.
+			\item We consider the case where Tx and Rx are using the S\&K strategy to communicate.
+			\item We record time series of transmissions at Tx and Rx, and using this we use Granger Causality to try to understand which direction the information is flowing.
+		\end{enumerate}
+	\item We discover that for reasonable choices of parameter values, the direction of Granger causal influence is opposite to the direction of information flow.
+\end{enumerate}
+
+\subsection{Survey of previous criticisms of Granger Causality}
+
+\begin{enumerate}
+	\item Granger causality vs.
+true causality - cite Pearl.
+Our argument is, in spirit, very similar to the common criticism of Granger causality, which states that GC is a statistical quantity.
+It does not answer the question of true causal influence: if X is turned off, will Y behave differenty?
+(Need to quote from Pearl here).However, in the situation we are studying, we ask a different question - we wish to understand whether or not GC gives the correct direction of *information flow*.
+In our analysis, causal influences are present in both directions.
+	\item Hidden node problem - hard to estimate GC correctly unless you are measuring all the nodes. (measurement problem)
+	\item Noisy measurements can change the direction - cite Friston, cite that economics paper
+\end{enumerate}
+
+Note: The technical objections presented above (apart from GC-is-not-true-causality) indicate that estimating GC is hard, because it is hard to measure all nodes, etc.
+There is some deficiency in the measurement process.
+If we could measure better (by sampling more nodes, or with lesser noise, or...) then we could get the correct answer.
+Our objection, on the other hand, is much more fundamental.
+*Even if* AR coeffs are computed correctly, and the GC metric is perfectly computed (as is the case in our counter-examples), GC would *still* not give the correct direction of information flow.
+To our knowledge, this is a completely novel argument against GC, and one that is far more severe than previous objections, at least as far as the usage of GC in neuroscience is concerned.
+
+Other objections we could add - Kun Zhang's paper (subsampling can cause errors), effective connectivity or plumbing? (venous drainage),
+
+\subsection{Possible criticisms of this work}
+\begin{enumerate}
+	\item Circularity - there is no circularity of argument.
+We are just assuming that there are sufficient number of trials to fully estimate the coefficients of the non-stationary AR process.
+We describe how this might be achieved in section ...
+(maybe this should come at the end?)
+	\item Math is not complicated - this is a strength of our analysis.
+Even simple counter-examples that do not employ difficult proof techniques are able to show that GC does not always correctly estimate the direction of info flow.
+\end{enumerate}
+
+\subsection{Shortcomings in our analysis}
+\begin{enumerate}
+	\item Not evolving source - not a stationary AR process.
+To circumvent this, we discuss estimation of non-stationary AR coeffs, since this is a neuroscience experiment, so we have many realizations of the same AR process in the form of multiple trials (cite Ding again.
+Quote if necessary) A future work will address this shortcoming by examining how GC works with an evolving source that produces a stationary process.
+	\item We restrict ourselves to Gaussian noise for simplicity of analysis.
+Noise in the actual brain is likely to be very different.
+However, AR processes are often used to model ERPs (need a citation for this).
+	\item In the S\&K algo, we have simplified power and energy constraints
+	\item This is for a simple two-node network. But for larger networks, the issue of incorrectly estimating the direction will probably only get exacerbated.
+\end{enumerate}
+
+\subsection{Discussion}
+\begin{enumerate}
+	\item Some of this misconception might have arisen because of Massey's directed info conservation - we expect that dir info in the forward dirn goes up if the other one goes down - but this is a misnomer.
+(Do we still need this?
+I am increasingly of the opinion that directed info / GC capture nothing about the message)
+\end{enumerate}
+
+\section{ghgh}
 

### 2015-07-06 paper.tex
Modify abstract.

@@
 
+% Abstract --------------------------------------------------------------------
+
 \begin{abstract}
-%\boldmath
-Not without controversy, Granger causality and, more recently, directed information, are two measures used to estimate ``causal influence'' of one stochastic process on another.
-Recently, many works interpret direction of causal influence as the direction of ``information flow.'' To test this latter interpretation, this paper uses two simple experiments to test whether these measures predict the correct direction of information flow, \textit{i.e.}, from the transmitting to the receiving node.
-These measures have been used extensively in neuroscience, and complex interconnected systems in general, to infer causal influences, and more recently, to infer direction of information flows.
-Using the problem of communication across feedback channels (first with noiseless feedback, and then with noisy feedback), and using a strategy inspired by the work of Schalkwijk and Kailath, we show that the direction of information flow can be opposite to that inferred using Granger causality and Directed Information.
-Thus, while it might be reasonable to infer direction of causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
+	Granger causality is an established measure of the ``causal influence'' that one stochastic process has on another.
+Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer causal influences.
+Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow'' {\color{red}(this is akin to, but different from, a criticism of Granger Causality that compares it to true causality, as defined by Judea Pearl)}.
+We test whether these measures correctly predict the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow is known by design.
+The experiments are based on a communication system with a feedback channel, upon which a strategy inspired by the work of Schalkwijk and Kailath is employed.
+We show that in these experiments, the direction of information flow can be opposite to that which is inferred by using Granger causality and Directed Information.
+Thus, while it might be reasonable to infer the direction of causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
 \end{abstract}

### 2015-07-06 paper.tex
Fix images and references.
Edit abstract; add motivation.

@@
   % driver is specified.
-  % \usepackage[dvips]{graphicx}
+  \usepackage[dvips]{graphicx}
   % declare the path(s) where your graphic files are
@@
   % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
+  \DeclareGraphicsExtensions{.eps}
 \fi
@@
 Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer causal influences.
-Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow'' {\color{red}(this is akin to, but different from, a criticism of Granger Causality that compares it to true causality, as defined by Judea Pearl)}.
+Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow'', which to us appears to be a questionable extrapolation {\color{red}(the question we raise is akin to, but different from, a criticism of Granger Causality that compares it to true causality, as defined by Judea Pearl)}.
 We test whether these measures correctly predict the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow is known by design.
 The experiments are based on a communication system with a feedback channel, upon which a strategy inspired by the work of Schalkwijk and Kailath is employed.
-We show that in these experiments, the direction of information flow can be opposite to that which is inferred by using Granger causality and Directed Information.
+We show that in these experiments, the direction of information flow can be opposite to that which is inferred using Granger causality and Directed Information.
 Thus, while it might be reasonable to infer the direction of causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
 \end{abstract}
+
 \IEEEpeerreviewmaketitle
+
+% Introduction ----------------------------------------------------------------
 
@@
 
+% Intro: Motivation -----------------------------------------------------------
+
 \subsection{Motivation}
+\label{subsec:motivation}
 
-\begin{enumerate}
-	\item Increased interest in understanding (a) functional connectivity maps and (b) computational networks of the brain.
-	\item Evidenced by funding - Brain Initiative, Human Brain project.
-	\item Quote one of the missions of the Brain Initiative - esp. for connectivity maps, but also for computation
-	\item Quote papers that try to understand how *information flows* in the brain. (past references from the ISIT paper - look for some more).
-	\item Last line: "The question we raise is - Do these directed connectivity maps, as determined by directional causal influence measures such as Granger causality, correctly identify the directions along which information flows in the brain?"
-\end{enumerate}
+This work is in large part motivated by a recent surge of interest in understanding neural circuits --- the connectivity and dynamic activity of parts of the brain --- and how they give rise to behavior and experience.
+This is evidenced in the launching of the BRAIN initiative in the US and the Human Brain Project in Europe.
+To quote from \emph{BRAIN 2025: A Scientific Vision}\footnote{The BRAIN Working Group's report to the Advisory Committee to the Director of the NIH}, we wish to ``map connected neurons in local circuits and distributed brain systems, enabling an understanding of the relationship between neuronal structure and function'', clearly indicating the move towards understanding (a) the connectivity and (b) the computational function of different brain regions.
+While the question of how the brain computes has been of immense interest for several decades, only recently have measurement techniques become sophisticated enough to be able to simultaneously record the activity of multiple neurons, or multiple neural populations.
+
+In an effort to make headway on these goals, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how information flows in the brain ({\color{red} ??}), or to acquire directed maps of functional connectivity ({\color{red} ??}).
+For instance, in~\cite{Brovelli2004BetaOscillations}, Granger causal influences that are measured between somatosensory and motor sites are said to ``support the idea that somatosensory feedback provides information to the sensorimotor system that is used to control motor output''.
+This raises the question: do these directed connectivity maps, as determined by directional causal influence measures such as Granger causality, correctly identify the directions along which information flows in the brain?
 
@@
 \bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv,MyMainBibliographyISIT15}
+\bibliography{IEEEabrv,references}
 

### 2015-07-06 paper.tex
Add section on GC used in neuroscience.

@@
 \documentclass[conference]{IEEEtran}
+%\documentclass[conference,draftcls]{IEEEtran}
 
@@
 \subsection{Motivation}
-\label{subsec:motivation}
+\label{sec:motivation}
 
@@
 
-\begin{enumerate}
-	\item We take Brovelli et.\ al.\ and Ding et.\ al.\ as the basis / canonical reference, as this appears to be generally accepted by the community (Cite review papers that cite these two papers; cite Anil Seth's MVGC package paper).
-	\item Explain Granger causality
-	\item Explain the exact method used - including trials, pre-processing steps, etc.\ - as described in the Bressler paper.
-	\item Granger's original analysis does not compare GC on forward and reverse links, however, this appears to be the commonly done in practice (cite Brovelli and review papers if possible)
-	\item State our problem statement again - we take issue not with any of these steps, but with the interpretation that if greater Granger causal influences are seen to be from X to Y, then information must be flowing from X to Y in the brain.
-\end{enumerate}
+We take~\cite{Brovelli2004BetaOscillations} and~\cite{Ding1999ShortWindow} to be the canonical references for the application of Granger causality to estimate causal influences, as they seem to be widely accepted by the community (\cite{Bressler2011WienerGranger},{\color{red} \cite{Roebroeck2005MappingDirected}},~\cite{Ding2006GrangerCausality}, \cite{Barnett2014MVGC}).
+Here, we briefly describe how Granger Causal influence is quantified, and how it is computed in these works.
 
-\subsection{Our counter-examples and main result}
+Granger causality, as originally described by Granger~\cite{Granger1969InvestigatingCausal}, measures the level of influence that one process $\{X\}$ has on another process $\{Y\}$.
+The analysis compares the \textit{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$.
+The Granger causality metric is the ratio of these errors.
+Many variations of Granger causality have also been considered, including a generalization --- the Directed Information (see~\cite{MasseyCausality,quinn,weissman}) --- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
 
-\begin{enumerate}
-	\item Define Granger causal influence and directed info in English again.
-Define information flow (use examples, or define formally?
-Formal definition should come in the math sections).
-Describe the problem again - is the direction of causal influence the same as the direction of information flow?
-	\item Describe the counter-examples in English.
-		\begin{enumerate}
-			\item Transmitter has the message (msg is known at the transmitter before it is known at the receiver - this is what distinguishes Tx from Rx).
-Therefore, the true direction in which information is flowing is *known* to us, by design of the experiment.
-This is the *ground truth*, against which we can compare the results of Granger causality.
-			\item Feedback link is present - we consider two cases: feedback link noiseless, and feedback link noisy.
-			\item We consider the case where Tx and Rx are using the S\&K strategy to communicate.
-			\item We record time series of transmissions at Tx and Rx, and using this we use Granger Causality to try to understand which direction the information is flowing.
-		\end{enumerate}
-	\item We discover that for reasonable choices of parameter values, the direction of Granger causal influence is opposite to the direction of information flow.
-\end{enumerate}
+Granger causality computation, in general, requires that the processes in consideration be stationary.
+In neuroscientific experiments, this is usually not the case, as the express objective of the experiment is to understand brain function while the brain is subjected to a task.
+In order to circumvent this problem, data analysis techniques either assume stationarity~\cite{Brovelli2004BetaOscillations}, or work with time-windows short enough that the processes are in fact stationary within those windows~\cite{Ding2006GrangerCausality}.
+In the latter case, Granger causal influences are computed for each window individually, in order to produce a dynamic description of brain connectivity as the brain carries out the said task.
 
-\subsection{Survey of previous criticisms of Granger Causality}
+In order to determine the direction of causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is \emph{compared} to that from $\{Y\}$ to $\{X\}$.
+We note here that Granger's original analysis does not compare this metric on forward and reverse links, however, this appears to be commonly done in practice.
+{\color{red} TODO: Cite Tsachy's paper here.}
+
+While we accept that Granger Causal influence can be accurately computed (provided measurements are taken suitably~\ref{sec:previous-criticisms-of-gc}) and can be useful in many situations~\ref{sec:discussion}, interpreting the direction of Granger Causal influence as the direction of information flow can be erroneous, as we demonstrate in this paper.
+
+\subsection{A short survey of previous criticisms of Granger Causality}
+
+Several objections have been leveled against Granger Causality in the past.
+We give, here, a short overview of these and describe why our objection is novel, and possibly more fundamental in nature, at least in the context of its usage in neuroscience.
 
@@
 Other objections we could add - Kun Zhang's paper (subsampling can cause errors), effective connectivity or plumbing? (venous drainage),
+
+\subsection{Our counter-examples and our main result}
+
+\begin{enumerate}
+	\item Define Granger causal influence and directed info in English again.
+Define information flow (use examples, or define formally?
+Formal definition should come in the math sections).
+Describe the problem again - is the direction of causal influence the same as the direction of information flow?
+	\item Describe the counter-examples in English.
+		\begin{enumerate}
+			\item Transmitter has the message (msg is known at the transmitter before it is known at the receiver - this is what distinguishes Tx from Rx).
+Therefore, the true direction in which information is flowing is *known* to us, by design of the experiment.
+This is the *ground truth*, against which we can compare the results of Granger causality.
+			\item Feedback link is present - we consider two cases: feedback link noiseless, and feedback link noisy.
+			\item We consider the case where Tx and Rx are using the S\&K strategy to communicate.
+			\item We record time series of transmissions at Tx and Rx, and using this we use Granger Causality to try to understand which direction the information is flowing.
+		\end{enumerate}
+	\item We discover that for reasonable choices of parameter values, the direction of Granger causal influence is opposite to the direction of information flow.
+\end{enumerate}
 

### 2015-07-06 paper.tex
Add section on previous criticisms of GC.

@@
 
-While we accept that Granger Causal influence can be accurately computed (provided measurements are taken suitably~\ref{sec:previous-criticisms-of-gc}) and can be useful in many situations~\ref{sec:discussion}, interpreting the direction of Granger Causal influence as the direction of information flow can be erroneous, as we demonstrate in this paper.
+While we accept that Granger Causal influence can be accurately computed (provided measurements are taken suitably; see Section~\ref{sec:previous-criticisms-of-gc}) and can be useful in many situations (see Section~\ref{sec:discussion}), interpreting the direction of Granger Causal influence as the direction of information flow can be erroneous, as we demonstrate in this paper.
 
@@
 \begin{enumerate}
-	\item Granger causality vs.
-true causality - cite Pearl.
-Our argument is, in spirit, very similar to the common criticism of Granger causality, which states that GC is a statistical quantity.
-It does not answer the question of true causal influence: if X is turned off, will Y behave differenty?
-(Need to quote from Pearl here).However, in the situation we are studying, we ask a different question - we wish to understand whether or not GC gives the correct direction of *information flow*.
-In our analysis, causal influences are present in both directions.
-	\item Hidden node problem - hard to estimate GC correctly unless you are measuring all the nodes. (measurement problem)
-	\item Noisy measurements can change the direction - cite Friston, cite that economics paper
+	\item First, Granger Causality suffers from what we call the ``hidden node problem''.
+If two observed nodes receive causal influences from a third, latent node, then a Causal influence may be detected between the observed nodes, even if they are independent of each other~\cite{PearlCausalitychapterasdfads}.
+All nodes need to be observed, therefore, to avoid finding spurious influences.
+	\item Second, if the measurements from each node are differentially affected by noise, then the predicted direction of causal influence might be opposite to the true direction of influence~\cite{Friston34234, EcoPaper3q4234}.
+Measurements need to be relatively noiseless and precise in order to obtain the correct direction of causal influece.
+	\item Third, subsampling the processes can produce incorrect results~\cite{KunZhangsPaper}. {\color{red} TODO: Write another line here - look at the paper for this.}
 \end{enumerate}
 
-Note: The technical objections presented above (apart from GC-is-not-true-causality) indicate that estimating GC is hard, because it is hard to measure all nodes, etc.
-There is some deficiency in the measurement process.
-If we could measure better (by sampling more nodes, or with lesser noise, or...) then we could get the correct answer.
+It is important to note that the technical objections listed above are all concerned with measurement.
+They indicate that incorrected Granger Causal influences may be computed if there is some deficiency in the measurement process.
+These objections can be resolved by taking better measurements (by sampling more nodes, with higher signal-to-noise ratio, or {\color{red} by not subsampling}).
+
 Our objection, on the other hand, is much more fundamental.
-*Even if* AR coeffs are computed correctly, and the GC metric is perfectly computed (as is the case in our counter-examples), GC would *still* not give the correct direction of information flow.
-To our knowledge, this is a completely novel argument against GC, and one that is far more severe than previous objections, at least as far as the usage of GC in neuroscience is concerned.
-
-Other objections we could add - Kun Zhang's paper (subsampling can cause errors), effective connectivity or plumbing? (venous drainage),
+\emph{Even if} the measurements are made with infinite accuracy, and the regression coefficients associated with computing Granger Causality are precisely estimated, and the Granger Causality metric is perfectly computed (as is the case in our counter-examples), Granger Causal influence would \emph{still} not give the correct direction of information flow.
+To our knowledge, this is a completely novel argument against Granger Causality, and one that is far more serious than previous objections, at least as far as its use in neuroscience is concerned.
 
@@
 
-\subsection{Possible criticisms of this work}
+\subsection{Possible objections to, and shortcomings of this work}
 \begin{enumerate}
+	\item Granger causality vs.
+true causality - cite Pearl.
+Our argument is, in spirit, very similar to the common criticism of Granger causality, which states that GC is a statistical quantity.
+It does not answer the question of true causal influence: if X is turned off, will Y behave differenty?
+(Need to quote from Pearl here).However, in the situation we are studying, we ask a different question - we wish to understand whether or not GC gives the correct direction of *information flow*.
+In our analysis, causal influences are present in both directions.
 	\item Circularity - there is no circularity of argument.
@@
 Even simple counter-examples that do not employ difficult proof techniques are able to show that GC does not always correctly estimate the direction of info flow.
-\end{enumerate}
-
-\subsection{Shortcomings in our analysis}
-\begin{enumerate}
 	\item Not evolving source - not a stationary AR process.

### 2015-07-06 paper.tex
Copy-paste section on counterexamples.

@@
 
+{\color{red} Needs re-write}
+
+This paper considers two experiments (introduced in Section~\ref{sec:sk}) where a transmitter $Tx$ wants to communicate a message to a receiver $Rx$ in presence of a feedback channel (in one experiment, the feedback link is noiseless, while in the other it is noisy).
+We assume that the experimenter is able to record the transmissions of $Tx$ and $Rx$ using some probing mechanism.
+Provided with these measurements, the experimenter wants to estimate which node is $Tx$ and which is $Rx$.
+Our results, derived in Section~\ref{sec:analytical-results} and numerically illustrated in Section~\ref{sec:numerical-results}, show that both Granger causality and Directed Information can incorrectly estimate which node is which.
+The first experiment is for the (unrealistic) case of communicating across noiseless feedback channels.
+The second experiment allows for noise in the feedback channel as well.
+In both cases, linear strategies inspired from the strategy of Schalkwijk and Kailath~\cite{S&K} are used.
+
+Our goal here is to bring out the point that whether Granger causality and Directed Information can be used to interpret direction of information flows is an issue that can be, and perhaps should be, considered using thought experiments on simple communication problems where \emph{information flow direction, and quantity, is already known}.
+If the prediction of Granger causality, or some other measure of information flow, matches with the known information flow, then that measure can be more confidently used in experiments.
+While our results strongly suggest that one needs to exercise care in interpreting Granger causality and Directed Information dominance as an indicator for the direction of information flow, there are several shortcomings that need to be addressed in order to understand the issue at depth.
+These shortcomings are discussed in detail in Section~\ref{sec:conclusions}.
+
 \begin{enumerate}
@@
 \subsection{Possible objections to, and shortcomings of this work}
+
+A review of a previous submission raised several objections to this work, which we seek to clarify here. Further, there are a few shortcomings to our analysis, which could be room for future work.
+
 \begin{enumerate}

### 2015-07-06 paper.tex
Add section of possible objections.

@@
 
-A review of a previous submission raised several objections to this work, which we seek to clarify here. Further, there are a few shortcomings to our analysis, which could be room for future work.
+A review of a previous submission raised some objections to this work, which we seek to rectify here.
+Further, our analysis has certain shortcomings, which we acknowledge here, and could be room for future work.
 
 \begin{enumerate}
-	\item Granger causality vs.
-true causality - cite Pearl.
-Our argument is, in spirit, very similar to the common criticism of Granger causality, which states that GC is a statistical quantity.
-It does not answer the question of true causal influence: if X is turned off, will Y behave differenty?
-(Need to quote from Pearl here).However, in the situation we are studying, we ask a different question - we wish to understand whether or not GC gives the correct direction of *information flow*.
-In our analysis, causal influences are present in both directions.
-	\item Circularity - there is no circularity of argument.
-We are just assuming that there are sufficient number of trials to fully estimate the coefficients of the non-stationary AR process.
-We describe how this might be achieved in section ...
-(maybe this should come at the end?)
-	\item Math is not complicated - this is a strength of our analysis.
-Even simple counter-examples that do not employ difficult proof techniques are able to show that GC does not always correctly estimate the direction of info flow.
-	\item Not evolving source - not a stationary AR process.
-To circumvent this, we discuss estimation of non-stationary AR coeffs, since this is a neuroscience experiment, so we have many realizations of the same AR process in the form of multiple trials (cite Ding again.
-Quote if necessary) A future work will address this shortcoming by examining how GC works with an evolving source that produces a stationary process.
-	\item We restrict ourselves to Gaussian noise for simplicity of analysis.
-Noise in the actual brain is likely to be very different.
-However, AR processes are often used to model ERPs (need a citation for this).
-	\item In the S\&K algo, we have simplified power and energy constraints
-	\item This is for a simple two-node network. But for larger networks, the issue of incorrectly estimating the direction will probably only get exacerbated.
+	\item Firstly, and most importantly, previous work has already shown that Granger Causality is not true causality (as defined by Judea Pearl~\cite{PearlCausalityBook}.
+Our argument is, in spirit, very similar to this common criticism of Granger causality, which states that Granger Causal influence is a statistical quantity and that it does not answer the question of true causal influence: if X is turned off, will Y behave differenty?
+{\color{red} (Need to quote from Pearl here)}.However, in the situation we are studying, we ask a different question - we wish to understand whether or not Granger Causality gives the correct direction of \emph{information flow}.
+In our analysis, causal influences are present in both directions, even in a ``true-causality'' sense.
+	\item It might appear that we make a circular argument while computing the Granger Causal influences in our counter-example, since we supply the model for the stochastic process and use the theoretical coeffiecients of the model directly to compute the Granger Causality metric.
+However, as we state in Section~\ref{sec:figure-it-out}, we assume that the regression coefficients of the Autoregressive model can be exactly estimated (even if the AR process is non-stationary), and discuss how this might be achieved with the help of multiple trials.
+{\color{red} (Move to conclusions?)}
+	\item Our analysis does not tackle an information source that evolves with time.
+Hence, our communication process (inspired from the scheme of Schalkwijk and Kailath~\cite{S&K}) is non-ergodic.
+Since Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process, we compute the obvious generalization of Granger causality to non-ergodic processes.
+Nevertheless, future work will address a situation with a linear dynamical system as the information source.
+	\item Our experiments restrict themselves to Gaussian noise for simplicity, but neural spiking and spike-rate models for spikes tend to be very different from those used here.
+This is a clear direction for future work.
+	\item The power and energy constraints are somewhat oversimplified to make the analysis simpler. This is for simplicity of exposition. A more general analysis is a trivial extension.
+	\item We consider a simple point-to-point network.
+In general networks, this issue could be even more complex.
+However, since this issue shows up even for the simplest network, we feel that the problem will only be exacerbated when the network is large.
+	\item We would also like to note that this paper is \emph{not} mathematically deep.
+We believe that this is a strength of our analysis.
+Even simple counter-examples that do not employ difficult proof techniques are able to show that Granger Causality does not always correctly estimate the direction of information flow.
+{\color{red} (Move this?)}
 \end{enumerate}

### 2015-07-06 paper.tex
Edit introduction. Add references.

@@
 
-This work is in large part motivated by a recent surge of interest in understanding neural circuits --- the connectivity and dynamic activity of parts of the brain --- and how they give rise to behavior and experience.
-This is evidenced in the launching of the BRAIN initiative in the US and the Human Brain Project in Europe.
+This work is in large part motivated by a recent surge of interest in understanding neural circuits -- the connectivity and dynamic activity of different regions of the brain -- and how they give rise to behavior and experience.
+This is evidenced by the launching of the BRAIN initiative in the US and the Human Brain Project in Europe.
 To quote from \emph{BRAIN 2025: A Scientific Vision}\footnote{The BRAIN Working Group's report to the Advisory Committee to the Director of the NIH}, we wish to ``map connected neurons in local circuits and distributed brain systems, enabling an understanding of the relationship between neuronal structure and function'', clearly indicating the move towards understanding (a) the connectivity and (b) the computational function of different brain regions.
@@
 
-In an effort to make headway on these goals, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how information flows in the brain ({\color{red} ??}), or to acquire directed maps of functional connectivity ({\color{red} ??}).
+In order to understand how the brain performs computations, it could be useful to first understand the directions of \emph{information flow} in various parts of the brain (e.g.~\cite{blinowska2004granger,dhamala2008analyzing,nolte2008robustly,korzeniewska2003determination,schippers2010mapping} etc.).
+In an effort to make headway on the goals of \emph{BRAIN 2025}, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how this information flows (eg.~\cite{Brovelli2004BetaOscillations,goebel2003investigating,deshpande2008effective}), or to acquire directed maps of functional connectivity (eg.~\cite{friston2013analysing,goebel2003investigating,deshpande2008effective}).
 For instance, in~\cite{Brovelli2004BetaOscillations}, Granger causal influences that are measured between somatosensory and motor sites are said to ``support the idea that somatosensory feedback provides information to the sensorimotor system that is used to control motor output''.
@@
 
+% Intro: How GC is used in neuroscience today ---------------------------------
+
 \subsection{How Granger Causality is used in Neuroscience today}
+\label{sec:gc-used-in-neuroscience}
 
-We take~\cite{Brovelli2004BetaOscillations} and~\cite{Ding1999ShortWindow} to be the canonical references for the application of Granger causality to estimate causal influences, as they seem to be widely accepted by the community (\cite{Bressler2011WienerGranger},{\color{red} \cite{Roebroeck2005MappingDirected}},~\cite{Ding2006GrangerCausality}, \cite{Barnett2014MVGC}).
+Several works have outlined the procedures involved in using Granger Causality to estimate causal influences in the brain (\cite{bernasconi2000bidirectional,Brovelli2004BetaOscillations,Ding2000ShortWindow,Roebroeck2005MappingDirected,Bressler2011WienerGranger,Barnett2014MVGC,Ding2006GrangerCausality}).
 Here, we briefly describe how Granger Causal influence is quantified, and how it is computed in these works.
@@
 Granger causality, as originally described by Granger~\cite{Granger1969InvestigatingCausal}, measures the level of influence that one process $\{X\}$ has on another process $\{Y\}$.
-The analysis compares the \textit{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$.
-The Granger causality metric is the ratio of these errors.
-Many variations of Granger causality have also been considered, including a generalization --- the Directed Information (see~\cite{MasseyCausality,quinn,weissman}) --- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
+The analysis compares the \emph{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$.
+The Granger causality metric is the ratio of these errors, encapsulating the innovations that the process $\{X\}$ causally supplies to the process $\{Y\}$.
+Many variations of Granger causality have also been considered, including a generalization -- Directed Information (see~\cite{MasseyCausality,quinn,weissman}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
+These form possible alternatives for estimating the direction of causal influence, but Directed Information is a generalization of many of these metrics~\cite{quinn}.
 
-Granger causality computation, in general, requires that the processes in consideration be stationary.
+The computation of Granger causality, in general, requires that the processes under consideration be stationary.
 In neuroscientific experiments, this is usually not the case, as the express objective of the experiment is to understand brain function while the brain is subjected to a task.
-In order to circumvent this problem, data analysis techniques either assume stationarity~\cite{Brovelli2004BetaOscillations}, or work with time-windows short enough that the processes are in fact stationary within those windows~\cite{Ding2006GrangerCausality}.
+Typically, data analysis techniques either assume stationarity regardless, or circumvent the problem by working with time-windows short enough that the processes \emph{are} in fact stationary within those windows~\cite{Ding2000ShortWindow}.
 In the latter case, Granger causal influences are computed for each window individually, in order to produce a dynamic description of brain connectivity as the brain carries out the said task.
+The Granger Causality metrics can be accurately computed even for short time windows, because neuroscientific experiments usually have multiple trials, i.e.\ multiple realizations of the same processes, allowing for fitting regression coefficients accurately~\cite{Ding2000ShortWindow}.
 
 In order to determine the direction of causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is \emph{compared} to that from $\{Y\}$ to $\{X\}$.
+The direction of causal influence is taken to be the direction with a greater Granger Causality metric.
+Further, this direction of causal influence is interpreted to be the direction of informationf flow -- the interpretation we question in this paper.
+We also note that many of these works use a spectral version of Granger Causality, that supplies this metric as a function of frequency.
+It then becomes possible to also determine the brain wave frequency at which these influences occur.
+However, we restrict our analysis to the simpler non-spectral version of Granger Causality, since it is sufficient to for the purpose of our arguments.
+
 We note here that Granger's original analysis does not compare this metric on forward and reverse links, however, this appears to be commonly done in practice.
@@
 
+% Intro: Previous criticisms of GC --------------------------------------------
+
 \subsection{A short survey of previous criticisms of Granger Causality}
+\label{sec:previous-criticisms-of-gc}
 
@@
 	\item First, Granger Causality suffers from what we call the ``hidden node problem''.
-If two observed nodes receive causal influences from a third, latent node, then a Causal influence may be detected between the observed nodes, even if they are independent of each other~\cite{PearlCausalitychapterasdfads}.
+If two observed nodes receive causal influences from a third, latent node, then a causal influence may be detected between the observed nodes, even if they are independent of each other~\cite{pearl2009causality-hidden-node}.
 All nodes need to be observed, therefore, to avoid finding spurious influences.
-	\item Second, if the measurements from each node are differentially affected by noise, then the predicted direction of causal influence might be opposite to the true direction of influence~\cite{Friston34234, EcoPaper3q4234}.
+	\item Second, if the measurements from each node are differentially affected by noise, then the predicted direction of causal influence might be opposite to the true direction (\cite{nalatore2007mitigating,andersson2005testing}).
 Measurements need to be relatively noiseless and precise in order to obtain the correct direction of causal influece.
-	\item Third, subsampling the processes can produce incorrect results~\cite{KunZhangsPaper}. {\color{red} TODO: Write another line here - look at the paper for this.}
+	\item Third, subsampling the processes can produce misleading Granger Causal relations~\cite{gong2015discovering}.
+Performing Granger Causal analysis on subsampled time series can lead one to miss the causal influence.
+If pre-processing involves subsampling, then this should be done with care.
 \end{enumerate}
 
-It is important to note that the technical objections listed above are all concerned with measurement.
-They indicate that incorrected Granger Causal influences may be computed if there is some deficiency in the measurement process.
-These objections can be resolved by taking better measurements (by sampling more nodes, with higher signal-to-noise ratio, or {\color{red} by not subsampling}).
+It is important to note that the technical objections listed above are all concerned with \emph{measurement}.
+They indicate that incorrect Granger Causal influences may be computed if there is some deficiency in the measurement procedure.
+These objections can be resolved by taking better measurements (by sampling more nodes, with higher signal-to-noise ratio, etc.).
 
@@
 \emph{Even if} the measurements are made with infinite accuracy, and the regression coefficients associated with computing Granger Causality are precisely estimated, and the Granger Causality metric is perfectly computed (as is the case in our counter-examples), Granger Causal influence would \emph{still} not give the correct direction of information flow.
-To our knowledge, this is a completely novel argument against Granger Causality, and one that is far more serious than previous objections, at least as far as its use in neuroscience is concerned.
+To our knowledge, this is a completely novel argument against Granger Causality, and one that is far more serious than previous objections, at least as far as its usefulness in understanding brain function is concerned.
+
+% Intro: Our counter-examples -------------------------------------------------
 
 \subsection{Our counter-examples and our main result}
-
-{\color{red} Needs re-write}
+\label{sec:our-counterexamples}
 
@@
 We assume that the experimenter is able to record the transmissions of $Tx$ and $Rx$ using some probing mechanism.
-Provided with these measurements, the experimenter wants to estimate which node is $Tx$ and which is $Rx$.
-Our results, derived in Section~\ref{sec:analytical-results} and numerically illustrated in Section~\ref{sec:numerical-results}, show that both Granger causality and Directed Information can incorrectly estimate which node is which.
-The first experiment is for the (unrealistic) case of communicating across noiseless feedback channels.
-The second experiment allows for noise in the feedback channel as well.
-In both cases, linear strategies inspired from the strategy of Schalkwijk and Kailath~\cite{S&K} are used.
+Provided with these measurements, the experimenter wants to estimate the direction of information flow, which in this context is the direction of flow of the message.
+Our results, derived in Section~\ref{sec:analytical-results} and numerically illustrated in Section~\ref{sec:numerical-results}, show that both Granger causality and Directed Information can incorrectly estimate the direction of information flow.
+The first experiment considers the (unrealistic) case of communication across noiseless feedback channels.
+The second experiment allows for noise in the feedback channel.
+In both cases, linear strategies inspired from the scheme of Schalkwijk and Kailath~\cite{S&K} are used.
 
 Our goal here is to bring out the point that whether Granger causality and Directed Information can be used to interpret direction of information flows is an issue that can be, and perhaps should be, considered using thought experiments on simple communication problems where \emph{information flow direction, and quantity, is already known}.
-If the prediction of Granger causality, or some other measure of information flow, matches with the known information flow, then that measure can be more confidently used in experiments.
+If the direction of causal influence yielded by Granger causality or some other similar measure were to match the known direction of information flow, then that measure can be more confidently used in experiments.
 While our results strongly suggest that one needs to exercise care in interpreting Granger causality and Directed Information dominance as an indicator for the direction of information flow, there are several shortcomings that need to be addressed in order to understand the issue at depth.
-These shortcomings are discussed in detail in Section~\ref{sec:conclusions}.
+These shortcomings are discussed in detail in Section~\ref{sec:objections-and-shortcomings}.
+
+% Intro: Possible objections and shortcomings ---------------------------------
+
+\subsection{Possible objections to, and shortcomings of this work}
+\label{sec:objections-and-shortcomings}
+
+A review of a previous submission raised some objections to this work, which we seek to clarify here.
+Further, our analysis has certain shortcomings, which we acknowledge.
+These could form the basis for future work.
 
 \begin{enumerate}
-	\item Define Granger causal influence and directed info in English again.
-Define information flow (use examples, or define formally?
-Formal definition should come in the math sections).
-Describe the problem again - is the direction of causal influence the same as the direction of information flow?
-	\item Describe the counter-examples in English.
-		\begin{enumerate}
-			\item Transmitter has the message (msg is known at the transmitter before it is known at the receiver - this is what distinguishes Tx from Rx).
-Therefore, the true direction in which information is flowing is *known* to us, by design of the experiment.
-This is the *ground truth*, against which we can compare the results of Granger causality.
-			\item Feedback link is present - we consider two cases: feedback link noiseless, and feedback link noisy.
-			\item We consider the case where Tx and Rx are using the S\&K strategy to communicate.
-			\item We record time series of transmissions at Tx and Rx, and using this we use Granger Causality to try to understand which direction the information is flowing.
-		\end{enumerate}
-	\item We discover that for reasonable choices of parameter values, the direction of Granger causal influence is opposite to the direction of information flow.
-\end{enumerate}
-
-\subsection{Possible objections to, and shortcomings of this work}
-
-A review of a previous submission raised some objections to this work, which we seek to rectify here.
-Further, our analysis has certain shortcomings, which we acknowledge here, and could be room for future work.
-
-\begin{enumerate}
-	\item Firstly, and most importantly, previous work has already shown that Granger Causality is not true causality (as defined by Judea Pearl~\cite{PearlCausalityBook}.
-Our argument is, in spirit, very similar to this common criticism of Granger causality, which states that Granger Causal influence is a statistical quantity and that it does not answer the question of true causal influence: if X is turned off, will Y behave differenty?
-{\color{red} (Need to quote from Pearl here)}.However, in the situation we are studying, we ask a different question - we wish to understand whether or not Granger Causality gives the correct direction of \emph{information flow}.
-In our analysis, causal influences are present in both directions, even in a ``true-causality'' sense.
+	\item Firstly, and most importantly, previous work has already shown that presence of a Granger Causal influence does not imply true causation.
+Judea Pearl classifies Granger Causality as stemming from a ``statistical'' model, rather than from a ``causal'' model~\cite{pearl2009causality-gc-stat}.
+A potential issue, therefore, might be the question the novelty of this work, given that Pearl's classification could be seen to clear up the misconception.
+While our argument is, in spirit, very similar to this criticism of Granger causality, in the situation we study, we ask a different question - we wish to understand whether or not Granger Causality gives the correct direction of \emph{information flow}.
+Further, we give a concrete counter-example in this context, which is very relevant in the modern neuroscience setting.
 	\item It might appear that we make a circular argument while computing the Granger Causal influences in our counter-example, since we supply the model for the stochastic process and use the theoretical coeffiecients of the model directly to compute the Granger Causality metric.
@@
 	\item The power and energy constraints are somewhat oversimplified to make the analysis simpler. This is for simplicity of exposition. A more general analysis is a trivial extension.
+	\item We have also restricted ourselves to analyzing linear feedback communication strategies.
+In the presence of noise in the feedback link, linear communication strategies are known to be sub-optimal~\cite{YoungHanKimPaper}.
+In order to make a water-tight argument, we would need to show that Granger Causality fails to correctly predict the direction of information flow, even when an optimal communication strategy -- which would have to be non-linear.
+This could be scope for future work.
+However, we do not expect results to change dramatically -- when the feedback link is impaired by noise of only very low variance, the noiseless case should make for a good approximation of the system, and results should degrade gracefully, if there is any degradation at all.
 	\item We consider a simple point-to-point network.
@@
 Even simple counter-examples that do not employ difficult proof techniques are able to show that Granger Causality does not always correctly estimate the direction of information flow.
-{\color{red} (Move this?)}
+{\color{red} (Move this, but where?)}
 \end{enumerate}
@@
 \begin{enumerate}
-	\item Some of this misconception might have arisen because of Massey's directed info conservation - we expect that dir info in the forward dirn goes up if the other one goes down - but this is a misnomer.
+	\item \color{red} Some of this misconception might have arisen because of Massey's directed info conservation - we expect that dir info in the forward dirn goes up if the other one goes down - but this is a misnomer.
 (Do we still need this?
@@
 In order to arrive at such an understanding of the brain, a natural method is to probe it to understand its functional connectivity\footnote{This is different from structural or anatomical connectivity in that links may sometimes be present, but not active.}.
-Analyses are then performed on the time series thus obtained from the probes, in order to say something about the functional connectivity --- which components influence each other, and which direction is this influence in.
+Analyses are then performed on the time series thus obtained from the probes, in order to say something about the functional connectivity -- which components influence each other, and which direction is this influence in.
 One commonly used analysis is based on the computation of Granger causality~\cite{Granger} with the measurements from each node.
@@
 The analysis compares the \textit{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$.
-Many variations of Granger causality have also been considered, including a generalization --- the Directed Information (see~\cite{MasseyCausality,quinn,weissman}) --- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
+Many variations of Granger causality have also been considered, including a generalization -- the Directed Information (see~\cite{MasseyCausality,quinn,weissman}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
 %So another strategy that could be used to infer causal influence of one process, e.g.

### 2015-07-06 paper.tex
Remove old introduction.

@@
 \end{enumerate}
-
-\section{ghgh}
-
-This work is in large part motivated by a recent surge of interest in understanding neural circuits --- the connectivity and dynamic activity of parts of the brain --- and how they give rise to behavior and experience\footnote{This is evidenced in the launching of the BRAIN initiative in the US and the Human Brain Project in Europe.}.
-The question of how the brain computes has been of immense interest for several decades, but only recently have measurement techniques become sophisticated enough to be able to simultaneously record the activity of multiple neurons, or multiple neural populations.
-
-In order to understand how the brain performs computations, it could be useful to understand first the direction of information flows  in various parts of the brain (e.g.~\cite{blinowska2004granger,dhamala2008analyzing,nolte2008robustly,korzeniewska2003determination,schippers2010mapping} etc.).
-For instance, what one might seek is a component-wise block diagram of the brain, where the brain is broken down into discrete computational units.
-In this picture, each computational unit receives information, processes it, and then passes it on to another unit.
-
-In order to arrive at such an understanding of the brain, a natural method is to probe it to understand its functional connectivity\footnote{This is different from structural or anatomical connectivity in that links may sometimes be present, but not active.}.
-Analyses are then performed on the time series thus obtained from the probes, in order to say something about the functional connectivity -- which components influence each other, and which direction is this influence in.
-One commonly used analysis is based on the computation of Granger causality~\cite{Granger} with the measurements from each node.
-The Granger causality between two processes measures the level of influence that one process $\{X\}$ has on another process $\{Y\}$.
-The analysis compares the \textit{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$.
-Many variations of Granger causality have also been considered, including a generalization -- the Directed Information (see~\cite{MasseyCausality,quinn,weissman}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
-%So another strategy that could be used to infer causal influence of one process, e.g.
-$\{X\}$, on another process $\{Y\}$ is to estimate and compare $I(\m{X}\to\m{Y})$ and $I(\m{Y}\to\m{X})$ (normalized by $m$) in the limit of large $m$.
-
-The question of when Granger causality is true causality has been hotly debated for quite some time (see, e.g.~\cite{jacobs1979difficulties} for some of the issues).
-Nevertheless, Granger causality is widely used for inferring causal influence in a wide variety of applications, ranging from biology to economics.
-In this work, we are more interested in a different issue on use of Granger causality (and Directed Information) that has received little attention: the interpretation of the dominant direction of causal influence as the direction of information flow.
-Quoting~\cite{nolte2008robustly} for instance, ``To understand interacting systems it is of fundamental importance to distinguish the \emph{driver from the recipient}, and hence to be able to estimate the direction of information flow.
-The direction can be estimated with a temporal argument: the driver is earlier than the recipient implying that the driver contains information about the future of the recipient not contained in the past of the recipient while the reverse is not the case'' (emphasis has been added by us).
-Is the dominant direction of directed information really the direction of information flow?
-Can we assign communication labels of ``transmitter'' and ``receiver'' along the direction of directed information?
-
-%Granger causality and Directed Information can, then, be used to estimate the direction of ``information flow'', and Granger causality often is.
-But in order to achieve an understanding of how the brain computes, it is necessary to think of the brain regions as communicating, as from a transmitter to a receiver.
-The authors of this paper believe that ascribing a direction to information flow in the brain, towards a goal of understanding how it computes, is meaningful only if there is also an implicit ``labelling'' of the brain regions being discussed as ``Trasmitter'' and ``Receiver''.
-
-%In this paper, we question whether one is justified in assigning labels of ``Transmitter'' and ``Receiver'', based only on the activity we record from probes? 
-%As one might anticipate, such an assignment can be a somewhat ambiguous for a system with feedback loops. 
-This paper considers two experiments (introduced in Section~\ref{sec:sk}) where a transmitter $Tx$ wants to communicate a message to a receiver $Rx$ in presence of a feedback channel (in one experiment, the feedback link is noiseless, while in the other it is noisy).
-We assume that the experimenter is able to record the transmissions of $Tx$ and $Rx$ using some probing mechanism.
-Provided with these measurements, the experimenter wants to estimate which node is $Tx$ and which is $Rx$.
-Our results, derived in Section~\ref{sec:analytical-results} and numerically illustrated in Section~\ref{sec:numerical-results}, show that both Granger causality and Directed Information can incorrectly estimate which node is which.
-The first experiment is for the (unrealistic) case of communicating across noiseless feedback channels.
-The second experiment allows for noise in the feedback channel as well.
-In both cases, linear strategies inspired from the strategy of Schalkwijk and Kailath~\cite{S&K} are used.
-
-Our goal here is to bring out the point that whether Granger causality and Directed Information can be used to interpret direction of information flows is an issue that can be, and perhaps should be, considered using thought experiments on simple communication problems where \textit{information flow direction, and quantity, is already known}.
-If the prediction of Granger causality, or some other measure of information flow, matches with the known information flow, then that measure can be more confidently used in experiments.
-While our results strongly suggest that one needs to exercise care in interpreting Granger causality and Directed Information dominance as an indicator for the direction of information flow, there are several shortcomings that need to be addressed in order to understand the issue at depth.
-These shortcomings are discussed in detail in Section~\ref{sec:conclusions}.
-
-%\begin{itemize}
-
-%\item Motivation: Recent interest in understanding neural circuits, to understand how computation is being done -- the BRAIN initiative in the US and the Blue Brain (?) project for mapping the brain in Europe.
-The ultimate goal is to understand how the brain works, and potentially to use that to engineer better circuits and systems (if we can quote from either of the projects, even better!).
-Another motivation for this work is probing a circuit to understand how computation is being performed.
-Or probing a social network?
-
-%\item Leads us to the tale of: What if a biologist came across a radio.
-How would they examine it?
-How will they probe it and based on the data, what conclusions will they draw?
-This has a circuit parallel as well.
-
-%\item The question of causality has been hotly debated elsewhere (see, e.g.
-??).
-More recently, causality has come to be interpreted as the direction of information flows.
-We need as many citations for this as possible.
-Some of them: Korzeniewska2003 and papers that cite it.
-The more papers we can cite with info-flow directions in their title, the better.
-
-%\item In communication parlance, information flows from the transmitter to the receiver.
-Feedback from the receiver can be viewed as information-flow as well.
-However, this flow could be of little intellectual interest [...
-why?...] %information flows not only from transmitter to receiver, but also on the feedback link in the reverse direction.
-Biological links, as well as circuits, tend to have substantial feedback, and thus it is of great interest to develop techniques to assign the ``transmitting'' and ``receiving'' nodes.
-
-%\item This paper performs some experiments on inferring the direction of information flows. To be able to ask the question of whether the information flow is predicted in the correct direction, we
-
-%\item The first experiment is for the case with noiseless feedback.
-In this highly asymmetric case, we observe that directed information and Granger causality can incorrectly predict the direction of forward and reverse links.
-
-%\end{itemize}
-
-%Possible shortcomings of our approach/objections to our approach:
-%\begin{itemize}
-%\item The source is not an ergodic process.
-However, Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process.
-So we compute the obvious generalization of Granger causality to non-ergodic processes.
-%\item In feedback communication channels, info flows not just from the transmitter to the receiver, but also from the receiver to the transmitter.
-So as long as the biologist does not make the leap from direction of information flow to assigning labels such as ``transmitter,'' and ``receiver,'' they will be fine.
-But if they do not assign these labels, how is computing direction of info flow helping them understand brain circuits?
-This is unclear to us.
-It could help them, for instance, to understand epileptic seizure propagation.
-But that provides little insight into computation (besides, that is really not ``info-flow'' at all!).
-%\item Directed information is computed for a Gaussian case. 
-%\item Neural models for spikes tend to be very different from those used here. 
-%\item Power and energy constraints are simplified.
-%\item We consider a simple point-to-point network. In general networks, this issue could be even more complex.
-%\item Granger's original analysis does not compare Granger causality on forward and reverse links.
-We are doing so, in part because it is often empirically done.
-Also, the ``conservation of directed information,'' an equality first derived by Massey~\cite{??}, could be viewed as suggesting that an increase in directed info in one direction must reduce it in the other direction.
-This is not true.
-For instance, making $X$ and $Y$ uncorrelated reduces directed info in both directions (and their sum) to zero.
-In other words, Massey's law is not a conservation law in the sense of conservations laws of energy, momentum, or mass, where the total energy/momentum/mass (\textit{i.e.}, the RHS) remains conserved (\textit{i.e.}, constant) regardless of the experiment performed on the entities.
-%\item More counterexamples need to be studied to get a better sense of when these tools are applicable and when they're not.
-%\item The work that looks at noisy versions of observed transmissions already creates a concern on use of info flows.
-Our work bridges that work with info theory.
-Instead of being a one-off result, that work points to a deeper issue with use of directed info in interpreting info flows.
-%\end{itemize}
 

### 2015-07-06 paper.tex
Make the pdf to specifications.
Perform some clean-up.

@@
-\documentclass[conference]{IEEEtran}
+\documentclass[letterpaper, 10pt, conference]{ieeeconf}
 %\documentclass[conference,draftcls]{IEEEtran}
+
+\overrideIEEEmargins
 
@@
 %
-\ifCLASSINFOpdf
-  \usepackage[pdftex]{graphicx}
+%\ifCLASSINFOpdf
+  %\usepackage[pdftex]{graphicx}
   % declare the path(s) where your graphic files are
@@
   % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
+%\else
   % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
@@
   \DeclareGraphicsExtensions{.eps}
-\fi
+%\fi
 % graphicx was written by David Carlisle and Sebastian Rahtz. It is
@@
 \end{equation*}
-where $n$ is the number of iterations of the Schalkwijk-Kailath algorithm. For a proof, refer appendix~\ref{app:appendix3}.
+where $n$ is the number of iterations of the Schalkwijk-Kailath algorithm. For a proof, refer appendix~\ref{app:dir-info-fwd-noiseless}.
 
@@
 \end{align*}
-For a derivation of this, see appendix~\ref{app:appendix1}.
+For a derivation of this, see appendix~\ref{app:dir-info-fwd-noisy}.
 
@@
 \end{align*}
-For a derivation of this, see appendix~\ref{app:appendix2}.
+For a derivation of this, see appendix~\ref{app:dir-info-rev-noisy}.
 
 % --------------------------------------------------------------------------- %
-
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals/conferences use top floats
-% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
-% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
-% command of the stfloats package.
 
@@
 
-%%%%%%%%%%%%%%%%%%
 There are several shortcomings of this work that need to be addressed for a more comprehensive understanding of the issue.
@@
 \section{Directed Information in the forward direction, noiseless feedback}
-\label{app:appendix3}
+\label{app:dir-info-fwd-noiseless}
 
@@
 \section{Directed Information in the forward direction, with noisy feedback}
-\label{app:appendix1}
+\label{app:dir-info-fwd-noisy}
 
@@
 
-% you can choose not to have a title for an appendix
-% if you want by leaving the argument blank
 \section{Directed Information in the reverse direction, with noisy feedback}
-\label{app:appendix2}
+\label{app:dir-info-rev-noisy}
 
@@
 	I(0*\widehat\Theta^{n-1} \rightarrow X^n) &= \sum_{i=0}^{n-1}{I(X_{i+1}; \widehat\Theta^i | X^i)} \\
-											  &= \sum_{i=0}^{n-1}{h(X_{i+1} | X^i) - h(X_{i+1} | X^i, \widehat\Theta^i)} \numberthis \label{eq:dir-info-rev-noisy}
+											  &= \sum_{i=0}^{n-1}{h(X_{i+1} | X^i) - h(X_{i+1} | X^i, \widehat\Theta^i)} \numberthis
 \end{align*}

### 2015-07-06 paper.tex
Add subsection on estimating AR coeffs.

@@
 Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer causal influences.
-Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow'', which to us appears to be a questionable extrapolation {\color{red}(the question we raise is akin to, but different from, a criticism of Granger Causality that compares it to true causality, as defined by Judea Pearl)}.
+Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow''.
+We ask: is the direction of causal influence, as predicted by Granger Causality and Directed Information always the same as the direction of information flow?
 We test whether these measures correctly predict the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow is known by design.
@@
 
-A review of a previous submission raised some objections to this work, which we seek to clarify here.
+A review of a submission to a previous conference raised some objections to this work, which we discuss to clarify our perspective.
 Further, our analysis has certain shortcomings, which we acknowledge.
-These could form the basis for future work.
+These will form the basis for future work.
 
@@
 	\item It might appear that we make a circular argument while computing the Granger Causal influences in our counter-example, since we supply the model for the stochastic process and use the theoretical coeffiecients of the model directly to compute the Granger Causality metric.
-However, as we state in Section~\ref{sec:figure-it-out}, we assume that the regression coefficients of the Autoregressive model can be exactly estimated (even if the AR process is non-stationary), and discuss how this might be achieved with the help of multiple trials.
+However, as we state in Section~\ref{sec:estimating-ar-coeffs}, we assume that the regression coefficients of the Autoregressive model can be exactly estimated (even if the AR process is non-stationary), and discuss how this might be achieved with the help of multiple trials.
 {\color{red} (Move to conclusions?)}
@@
 
+\subsection{A note on non-stationarity and estimating regression coefficients}
+\label{sec:estimating-ar-coeffs}
+
+The Granger Causality and Directed Inforamtion metrics are a function of the regression coefficients estimated from the data by fitting the models given by equations~\ref{eq:ar-model-x-x} and~\ref{eq:ar-model-x-x-theta}.
+In our analysis, we described the data-generation model: the algorithm inspired by the Schalkwijk and Kailath scheme for feedback communication.
+We then proceeded to use the regression coefficients generated by this model directly.
+This could be construed as erroneous: we ought to simulate the data generation, and compute the regression coefficients from the generated data.
+This would better model the actions of the neuroscientist who seeks to perform a Granger Causal analysis.
+
+We justify our use of the theoretical regression coefficients as follows: we assume that the regression coefficients can be accurately estimated from data, since neuroscientific experiments typically record the \emph{same} processes multiple times -- these are called ``trials''.
+The availability of multiple trials of the same process can be leveraged to estimate the regression coefficients accurately, even if the processes are non-stationary.
+
+Suppose we record two non-stationary processes, $\{X_t\}_{t=1}^n$ and $\{Y_t\}_{t=1}^n$, for which we seek to compute the Granger Causality metrics.
+To this end, we must find coefficients $\alpha_j(t)$ and $\beta_j(t)$ to minimize the error in fitting the models given by equations~\ref{eq:ar-model-x-x} and~\ref{eq:ar-model-x-x-theta}.
+Note that $\alpha$ and $\beta$ depend on $t$, since we assume the process is non-stationary.
+However, since we record the \emph{same} process in each trial, the $\alpha_j(t)$ and $\beta_j(t)$ are constant across trials.
+Estimating them from data that has many trials is then a simple matter of linear regression.
+
+For a given time instant $t$, the $i$\textsuperscript{th} trial is modeled as
+\begin{equation*}
+	X_t^{(i)} = \sum_{j=1}^p \alpha_j(t) X_{t-j}^{(i)} + \epsilon_t^{(i)}
+\end{equation*}
+Note that $\alpha_j(t)$ does not depend on $i$. Collecting variables across $N$ trials, we can write the full model in vector form:
+\begin{align*}
+	\left[ \begin{array}{c}
+			X_t^{(1)} \\
+			X_t^{(2)} \\
+			\vdots \\
+			X_t^{(N)}
+	\end{array} \right]
+	&=
+	\left[ \begin{array}{cccc}
+			X_{t-1}^{(1)} & X_{t-2}^{(1)} & \cdots & X_{t-p}^{(1)} \\
+			X_{t-1}^{(2)} & X_{t-2}^{(2)} & \cdots & X_{t-p}^{(2)} \\
+			\vdots &  & \ddots & \vdots \\
+			X_{t-1}^{(N)} & X_{t-2}^{(N)} & \cdots & X_{t-p}^{(N)} \\
+	\end{array} \right]
+	\left[ \begin{array}{c}
+			\alpha_1(t) \\
+			\alpha_2(t) \\
+			\vdots \\
+			\alpha_p(t)
+	\end{array} \right] \\
+	&\qquad \qquad \qquad +
+	\left[ \begin{array}{c}
+			\epsilon_t^{(1)} \\
+			\epsilon_t^{(2)} \\
+			\vdots \\
+			\epsilon_t^{(N)}
+	\end{array} \right]
+\end{align*}
+
+If we call the vector on the LHS $\bf{Y}$ and the matrix on the RHS $\bf{X}$, then the vector of $\alpha_j(t)$'s ($\bf{\alpha}(t)$) can be estimated at time instant $t$ using Ordinary Least Squares:
+\begin{equation*}
+\hat{\bf{\alpha}}(t) = (\bf{X}^T \bf{X})^{-1} \bf{X}^T \bf{Y}
+\end{equation*}
+This is an unbiased and consistent estimator for $\bf{\alpha}(t)$.
+This analysis can be trivially extended to the model described by equation~\ref{eq:ar-model-x-x-theta}.
+With a sufficiently large number of trials, therefore, the regression coefficients can be estimated to arbitrarily high accuracy.
+
+As a final point, we note that estimating the regression coefficients accurately is a form of worst-case analysis.
+As mentioned at the end of section~\ref{sec:previous-criticisms-of-gc}, we see that \emph{despite} being computed accurately, the metrics of Granger Causality and Directed Information incorrectly estimate the direction of information flow.
+
 % --------------------------------------------------------------------------- %

### 2015-07-06 paper.tex
Fix missing labels and refs.

@@
 
-While we accept that Granger Causal influence can be accurately computed (provided measurements are taken suitably; see Section~\ref{sec:previous-criticisms-of-gc}) and can be useful in many situations (see Section~\ref{sec:discussion}), interpreting the direction of Granger Causal influence as the direction of information flow can be erroneous, as we demonstrate in this paper.
+While we accept that Granger Causal influence can be accurately computed (provided measurements are taken suitably; see Section~\ref{sec:previous-criticisms-of-gc}) and can be useful in many situations (see Section~\ref{sec:conclusions}), interpreting the direction of Granger Causal influence as the direction of information flow can be erroneous, as we demonstrate in this paper.
 
@@
 \begin{equation*}
-	X_i = \sum_{j=1}^{p}{\alpha_j X_{i-j}} + \epsilon_i
+	X_i = \sum_{j=1}^{p}{\alpha_j X_{i-j}} + \epsilon_i \numberthis \label{eq:ar-model-x-x}
 \end{equation*}
@@
 \begin{equation*}
-	X_i = \sum_{j=1}^{p}{\alpha_j X_{i-j}} + \sum_{j=1}^{p}{\beta_j \widehat\Theta_{i-j}} + \tilde\epsilon_i
+	X_i = \sum_{j=1}^{p}{\alpha_j X_{i-j}} + \sum_{j=1}^{p}{\beta_j \widehat\Theta_{i-j}} + \tilde\epsilon_i \numberthis \label{eq:ar-model-x-x-theta}
 \end{equation*}

### 2015-07-06 paper.tex
Move the note on mathematical simplicity.

@@
 
+We find it interesting to note that the mathematical machinery used in this paper amounts to routine arithmetic.
+Even simple counter-examples that do not employ difficult proof techniques are able to demonstrate our main result.
+The authors wonder if more problems such as this exist, where mathematically simple information-theoretic counter-examples can lead to interesting conceptual results.
+
 % Intro: Possible objections and shortcomings ---------------------------------
@@
 However, as we state in Section~\ref{sec:estimating-ar-coeffs}, we assume that the regression coefficients of the Autoregressive model can be exactly estimated (even if the AR process is non-stationary), and discuss how this might be achieved with the help of multiple trials.
-{\color{red} (Move to conclusions?)}
+{\color{red} (Move to conclusions)}
 	\item Our analysis does not tackle an information source that evolves with time.
@@
 However, since this issue shows up even for the simplest network, we feel that the problem will only be exacerbated when the network is large.
-	\item We would also like to note that this paper is \emph{not} mathematically deep.
-We believe that this is a strength of our analysis.
-Even simple counter-examples that do not employ difficult proof techniques are able to show that Granger Causality does not always correctly estimate the direction of information flow.
-{\color{red} (Move this, but where?)}
 \end{enumerate}

### 2015-07-07 paper.tex
Make misc. edits; edit numerical results section.

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Some thought experiments on the applicability of Granger causality and Directed Information in inferring direction of information flows}
-\author{Praveen Venkatesh and Pulkit Grover\\Carnegie Mellon University}
+\title{Is the direction of Granger causal influence the same as the direction of information flow?}
+\author{Praveen Venkatesh\\vpraveen@cmu.edu\\Carnegie Mellon University}
+\author{Pulkit Grover\\pulkit@cmu.edu\\Carnegie Mellon University}
 
@@
 We ask: is the direction of causal influence, as predicted by Granger Causality and Directed Information always the same as the direction of information flow?
-We test whether these measures correctly predict the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow is known by design.
+We test whether these measures correctly predict the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow (the ``ground truth'') is known by design.
 The experiments are based on a communication system with a feedback channel, upon which a strategy inspired by the work of Schalkwijk and Kailath is employed.
@@
 
-We note here that Granger's original analysis does not compare this metric on forward and reverse links, however, this appears to be commonly done in practice.
-{\color{red} TODO: Cite Tsachy's paper here.}
+We note here that Granger's original analysis does not compare this metric on forward and reverse links, however, this is commonly done in practice even in areas beyond neuroscience (eg.~\cite{Brovelli2004BetaOscillations,jiao2013universal}).
 
@@
 
-It is important to note that the technical objections listed above are all concerned with \emph{measurement}.
-They indicate that incorrect Granger Causal influences may be computed if there is some deficiency in the measurement procedure.
+It is important to note that the technical objections listed above are all deficiencies in or limitations of \emph{measurement}.
+They indicate that incorrect Granger Causal influences may be computed if there is some deficiency or limitation in the measurement procedure.
 These objections can be resolved by taking better measurements (by sampling more nodes, with higher signal-to-noise ratio, etc.).
 
-Our objection, on the other hand, is much more fundamental.
-\emph{Even if} the measurements are made with infinite accuracy, and the regression coefficients associated with computing Granger Causality are precisely estimated, and the Granger Causality metric is perfectly computed (as is the case in our counter-examples), Granger Causal influence would \emph{still} not give the correct direction of information flow.
-To our knowledge, this is a completely novel argument against Granger Causality, and one that is far more serious than previous objections, at least as far as its usefulness in understanding brain function is concerned.
+Our objection, on the other hand, is more fundamental.
+\emph{Even if} the measurements are made with infinite accuracy, and the regression coefficients associated with computing Granger Causality are precisely estimated, and the Granger Causality metric is perfectly computed (as is the case in our counter-examples), Granger Causal influence may \emph{still} not yield the correct direction of information flow.
+To our knowledge, the argument that Granger causal influence can be opposite to the direction of information flow is a novel one.
+We believe that this argument is much more serious than previous objections, at least in the context of determining the directions of information flow in neuroscientific experiments, towards understanding the computational functions of brain regions.
 
@@
 
-\subsection{Discussion}
-\begin{enumerate}
-	\item \color{red} Some of this misconception might have arisen because of Massey's directed info conservation - we expect that dir info in the forward dirn goes up if the other one goes down - but this is a misnomer.
-(Do we still need this?
-I am increasingly of the opinion that directed info / GC capture nothing about the message)
-\end{enumerate}
-
 %Notes:
-
 %You only seem to be assuming that the variance is $\sigma_\theta^2$.
@@
 \subsection{Granger causality for the noiseless feedback case}
+\label{sec:gc-noiseless}
 
@@
 \subsection{Directed Information for the noiseless feedback case}
+\label{sec:dir-info-noiseless}
 
@@
 
-\section{Numerical results}
+\section{Numerical results demonstrating that Granger Causal influence can be opposite in direction to information flow}
 \label{sec:numerical-results}
-In the noiseless feedback case, the directed information on the forward link is finite, while that on the feedback link is infinite.
-However, the situation is more interesting when there is noise in the feedback link, as is illustrated in Fig.~\ref{fig:numerical} through numerical calculations of expressions in the last section.
+In the noiseless feedback case, the directed information on the forward link is finite, while that on the feedback link is infinite (refer sections~\ref{sec:gc-noiseless} and~\ref{sec:dir-info-noiseless}).
+However, for completeness, we examine the case when noise is present in the feedback link, as is illustrated in Fig.~\ref{fig:numerical}, through numerical calculation of expressions in the last section.
+For cases where the noise variance of the feedback link ($\sigma_R^2$) is moderately smaller than feedforward noise variance ($\sigma_N^2$), we observe that the Directed Information in the direction of the reverse link can dominate that in the direction of the forward link.
+With sufficiently many (albeit sometimes a large number of) iterations, the Directed Information in forward direction starts to dominate.
+However, the point at which this happens depends on the (often unknown) ratio of noise in these links.
 
@@
 In other words, Massey's law is not a conservation law in the sense of conservation laws of energy, momentum, or mass, where the total energy/momentum/mass (\textit{i.e.}, the RHS) remains conserved (\textit{i.e.}, constant) regardless of the experiment performed on the entities.
+
+% TODO: Include in final version:
+% Some of this misconception might have arisen because of Massey's directed info conservation - we expect that dir info in the forward dirn goes up if the other one goes down - but this is a misnomer.
+(Do we still need this?
+I am increasingly of the opinion that directed info / GC capture nothing about the message)
 

### 2015-07-07 paper.tex
Add conclusions and discussions section.

@@
 Further, we give a concrete counter-example in this context, which is very relevant in the modern neuroscience setting.
-	\item It might appear that we make a circular argument while computing the Granger Causal influences in our counter-example, since we supply the model for the stochastic process and use the theoretical coeffiecients of the model directly to compute the Granger Causality metric.
-However, as we state in Section~\ref{sec:estimating-ar-coeffs}, we assume that the regression coefficients of the Autoregressive model can be exactly estimated (even if the AR process is non-stationary), and discuss how this might be achieved with the help of multiple trials.
-{\color{red} (Move to conclusions)}
 	\item Our analysis does not tackle an information source that evolves with time.
@@
 
-There are several shortcomings of this work that need to be addressed for a more comprehensive understanding of the issue.
-First, our information source does not evolve with time, and hence our communication process (inspired from the scheme of Schalkwijk and Kailath~\cite{S&K} is non-ergodic.
-Since Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process, we compute the obvious generalization of Granger causality to non-ergodic processes.
-Nevertheless, future work will address a situation with a linear dynamical system as the information source.
-Second, our experiments restrict themselves to Gaussian noise for simplicity, but neural spiking and spike-rate models for spikes tend to be very different from those used here.
-This is a clear direction for future work.
-Third, the power and energy constraints are somewhat oversimplified to make the analysis simpler.
-This is for simplicity of exposition.
-A more general analysis is a trivial extension.
-Finally, we consider a simple point-to-point network.
-In general networks, this issue could be even more complex.
-However, since this issue shows up even for the simplest network, we feel that the problem will only be exacerbated when the network is large.
+We {\color{red} show(?)} that causal influence and information flow are distinct quantities.
+We demonstrate, by means of a concrete counter-example, that the direction predicted by causal influence metrics such as Granger Causality and Directed Information can be opposite to the true direction of information flow.
+There are, however, several shortcomings to our analysis, which we list in section~\ref{sec:objections-and-shortcomings}.
+We seek to address many of these shortcomings in future work.
 
-We note that Granger's original analysis does not compare Granger causality on forward and reverse links.
+It might appear that we make a circular argument while computing the Granger Causal influences in our counter-example, since we supply the model for the stochastic process and use the system parameters of the model directly to compute the Granger Causality metric.
+However, as we state in Section~\ref{sec:estimating-ar-coeffs}, we assume that the regression coefficients of the Autoregressive model can be exactly estimated (even if the AR process is non-stationary), and discuss how this might be achieved with the help of multiple trials.
+
+As a final remark, we emphasize that this work only demonstrates the error in \emph{interpreting} the direction of causal influence as the direction of \emph{information flow}.
+We do \emph{not} seek to invalidate much of the neuroscientific work that has been done in this direction; we merely caution against making {\color{red} (what might be construed as hopeful)} extrapolations from causal influences as to the manner in which information flows.
+
+We do not seek to understate the importance of determining causal influences in the brain; understanding causal influence itself may have a great deal of benefit.
+For instance, we might seek to understand the spread of activity in the brain during an epileptic siezure -- here, we are not concerned with how information is being transferred through the neural circuitry; we only seek to determine the source of the activity for the purpose of surgical intervention.
+
+%We note that Granger's original analysis does not compare Granger causality on forward and reverse links.
 We are doing so, in part because it is often done in practice.

### 2015-07-07 paper.tex
Correct spellings.

@@
 \title{Is the direction of Granger causal influence the same as the direction of information flow?}
-\author{Praveen Venkatesh\\vpraveen@cmu.edu\\Carnegie Mellon University}
-\author{Pulkit Grover\\pulkit@cmu.edu\\Carnegie Mellon University}
+\author{Praveen Venkatesh and Pulkit Grover\\Carnegie Mellon University}
 
@@
 The direction of causal influence is taken to be the direction with a greater Granger Causality metric.
-Further, this direction of causal influence is interpreted to be the direction of informationf flow -- the interpretation we question in this paper.
+Further, this direction of causal influence is interpreted to be the direction of information flow -- the interpretation we question in this paper.
 We also note that many of these works use a spectral version of Granger Causality, that supplies this metric as a function of frequency.
@@
 	\item Second, if the measurements from each node are differentially affected by noise, then the predicted direction of causal influence might be opposite to the true direction (\cite{nalatore2007mitigating,andersson2005testing}).
-Measurements need to be relatively noiseless and precise in order to obtain the correct direction of causal influece.
+Measurements need to be relatively noiseless and precise in order to obtain the correct direction of causal influence.
 	\item Third, subsampling the processes can produce misleading Granger Causal relations~\cite{gong2015discovering}.
@@
 
-The Granger Causality and Directed Inforamtion metrics are a function of the regression coefficients estimated from the data by fitting the models given by equations~\ref{eq:ar-model-x-x} and~\ref{eq:ar-model-x-x-theta}.
+The Granger Causality and Directed Information metrics are a function of the regression coefficients estimated from the data by fitting the models given by equations~\ref{eq:ar-model-x-x} and~\ref{eq:ar-model-x-x-theta}.
 In our analysis, we described the data-generation model: the algorithm inspired by the Schalkwijk and Kailath scheme for feedback communication.

### 2015-07-07 paper.tex
Pulkit's final edits.

@@
 Even simple counter-examples that do not employ difficult proof techniques are able to demonstrate our main result.
-The authors wonder if more problems such as this exist, where mathematically simple information-theoretic counter-examples can lead to interesting conceptual results.
+This simplicity leads us to think that this counter-example is not very special, and that directions of stronger Granger Causal influence and information flow might have little to do with each other in more complex and/or noisy networks.
 
@@
 
-We {\color{red} show(?)} that causal influence and information flow are distinct quantities.
 We demonstrate, by means of a concrete counter-example, that the direction predicted by causal influence metrics such as Granger Causality and Directed Information can be opposite to the true direction of information flow.
@@
 We do not seek to understate the importance of determining causal influences in the brain; understanding causal influence itself may have a great deal of benefit.
-For instance, we might seek to understand the spread of activity in the brain during an epileptic siezure -- here, we are not concerned with how information is being transferred through the neural circuitry; we only seek to determine the source of the activity for the purpose of surgical intervention.
+For instance, we might seek to understand the spread of activity in the brain during an epileptic siezure -- in such applications, we are not concerned with how information is being transferred through the neural circuitry; we only seek to determine the source of the activity for the purpose of surgical intervention.
 

### 2015-07-07 paper.tex
Clean up the latex file.

@@
 \documentclass[letterpaper, 10pt, conference]{ieeeconf}
-%\documentclass[conference,draftcls]{IEEEtran}
 
@@
 
-% *** CITATION PACKAGES ***
-%
-%\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-%\ifCLASSINFOpdf
-  %\usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-%\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
 \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
 \DeclareGraphicsExtensions{.eps}
-%\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
 
@@
 % *** MATH PACKAGES ***
-%
 \usepackage[cmex10]{amsmath}
 \usepackage{amssymb}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
 \interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
 
@@
 
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-%\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
 \hyphenation{op-tical net-works semi-conduc-tor}
@@
 \begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
 \title{Is the direction of Granger causal influence the same as the direction of information flow?}
@@
 
-% make the title area
 \maketitle
@@
 
-% --------------------------------------------------------------------------- %
+% Schalkwijk and Kailath scheme description -----------------------------------
 
@@
 In both cases, the estimators will be shown to be unbiased and consistent (the error mean is zero, and the error variance converges to zero).
+
+% S&K: Noiseless feedback -----------------------------------------------------
 
@@
 \end{align*}
+
+% S&K: Noisy feedback ---------------------------------------------------------
 
@@
 
+% Computing GC and Directed Information ---------------------------------------
+
 \section{Granger causality and directed information analyses for the strategies in Section~\ref{sec:sk}}
 \label{sec:analytical-results}
+
+% Computing GC: Noiseless feedback --------------------------------------------
 
@@
 
+% Computing Directed Information: Noiseless feedback --------------------------
+
 \subsection{Directed Information for the noiseless feedback case}
@@
 
+% Computing Directed Information: noisy feedback ------------------------------
+
 \subsection{Directed Information for the noisy feedback scenario}
@@
 
-\subsection{A note on non-stationarity and estimating regression coefficients}
+% Computing GC and DI: Estimating regression coefficients ---------------------
+
+\subsection{A note on estimating regression coefficients}
 \label{sec:estimating-ar-coeffs}
@@
 
-% --------------------------------------------------------------------------- %
+% Numerical results -----------------------------------------------------------
 
@@
 
+% Conclusions and discussions -------------------------------------------------
+
 \section{Conclusions and discussions}
@@
 As a final remark, we emphasize that this work only demonstrates the error in \emph{interpreting} the direction of causal influence as the direction of \emph{information flow}.
-We do \emph{not} seek to invalidate much of the neuroscientific work that has been done in this direction; we merely caution against making {\color{red} (what might be construed as hopeful)} extrapolations from causal influences as to the manner in which information flows.
+We do \emph{not} seek to invalidate much of the neuroscientific work that has been done in this direction; we merely caution against making (what might be construed as hopeful) extrapolations from causal influences as to the manner in which information flows.
 
@@
 I am increasingly of the opinion that directed info / GC capture nothing about the message)
+
+% Acknowledgements ------------------------------------------------------------
 

### 2015-07-07 paper.tex
Make corrections to the estimating coeffs section.

@@
 In our analysis, we described the data-generation model: the algorithm inspired by the Schalkwijk and Kailath scheme for feedback communication.
-We then proceeded to use the regression coefficients generated by this model directly.
-This could be construed as erroneous: we ought to simulate the data generation, and compute the regression coefficients from the generated data.
-This would better model the actions of the neuroscientist who seeks to perform a Granger Causal analysis.
+We then proceeded to use the system parameters of this model directly as regression coefficients in our Granger Causality computation.
+This could be construed as being erroneous: we ought to simulate the data generation, and estimate the regression coefficients from the generated data.
+This would better model the actions of the neuroscientist who seeks to perform Granger Causality analysis.
 
-We justify our use of the theoretical regression coefficients as follows: we assume that the regression coefficients can be accurately estimated from data, since neuroscientific experiments typically record the \emph{same} processes multiple times -- these are called ``trials''.
-The availability of multiple trials of the same process can be leveraged to estimate the regression coefficients accurately, even if the processes are non-stationary.
+We justify our knowledge of the system parameters and their use as regression coefficients in the following manner: we assume that the regression coefficients can be accurately estimated from data, since neuroscientific experiments typically record the \emph{same} processes multiple times -- these are called ``trials''.
+The availability of multiple trials of the same process can be leveraged to estimate the system parameters accurately, even if the processes are non-stationary.
 
@@
 This analysis can be trivially extended to the model described by equation~\ref{eq:ar-model-x-x-theta}.
-With a sufficiently large number of trials, therefore, the regression coefficients can be estimated to arbitrarily high accuracy.
+With a sufficiently large number of trials, therefore, the system parameters (to be used as regression coefficients in the Granger Causality analysis) can be estimated to arbitrarily high accuracy.
 
-As a final point, we note that estimating the regression coefficients accurately is a form of worst-case analysis.
+As a final remark, we note that estimating the regression coefficients accurately is a conservative assumption on our part.
 As mentioned at the end of section~\ref{sec:previous-criticisms-of-gc}, we see that \emph{despite} being computed accurately, the metrics of Granger Causality and Directed Information incorrectly estimate the direction of information flow.
+A rigorous analysis would warrant the computation of these coefficients in simulation, for a finite number of trials.
+It is our belief, however, that our result is unlikely to degrade if the coefficients are not estimated perfectly.
+Future work will address this matter in greater depth.
 

### 2015-07-07 paper.tex
Add SK block diagram figure.

@@
 \label{sec:sk}
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+	\centering
+	\includegraphics[width=3.2in]{sk-block-diagram}
+	\caption{A block-diagram representation of the communication system, describing the feedback channels and supplying notation for the variables used throughout the paper.}
+	\label{fig:sk-block-diagram}
+\end{figure}
+
 This section summarizes the analysis in the work of Schalkwijk and Kailath~\cite{S&K}, and a simple (and previously known) extension to a noisy feedback case.
@@
 Suppose we try to use the Granger causality measure of directed influence to infer the direction of information flow.
-As discussed in Section~\ref{sec:intro}, one might expect the direction of information flow --- and hence the direction of greater Granger causality --- to be from the transmitter to the receiver.
+As discussed in Section~\ref{sec:intro}, one might expect the direction of information flow -- and hence the direction of greater Granger causality -- to be from the transmitter to the receiver.
 

### 2015-07-07 paper.tex
Edit SK diagram caption.

@@
 	\includegraphics[width=3.2in]{sk-block-diagram}
-	\caption{A block-diagram representation of the communication system, describing the feedback channels and supplying notation for the variables used throughout the paper.}
+	\caption{A block-diagram representation of the communication system, describing the feedback channels and supplying notation for the variables used throughout the paper.
+Note that this diagram is the more general of the two cases discussed in sub-sections~\ref{sec:sk-noiseless} and~\ref{sec:sk-noisy}, as it contains noise in the feedback link.
+The former, noiseless, case is equivalent to setting $\sigma_R^2$ to 0.}
 	\label{fig:sk-block-diagram}
@@
 \subsection{Noiseless feedback: the Schalkwijk-Kailath strategy}
+\label{sec:sk-noiseless}
+
  In the first step, the transmitter sends\footnote{We assume that the power constraints are such that the scaling constant $\alpha$ in~\cite{S&K} is $1$.} $X_1 = \Theta$, which the receiver receives with added noise.
@@
 \subsection{Noisy feedback}
+\label{sec:sk-noisy}
+
 In the presence of noise in the feedback link, restricting our attention to linear strategies, we can use a simple modification of the Schalkwijk-Kailath strategy, incorporating the feedback\footnote{We note that implicitly, this strategy assumes an energy/SNR constraint on the feedback link.

### 2015-07-07 paper.tex
Rewrite parts of the math sections.

@@
 
-Therefore, in general, the transmitter sends $X_i = \Theta - \widehat\Theta_{i-1}$ and the receiver receives $Y_i = X_i + N_i$, where $N_i \sim \mathcal{N}(0, \sigma_N^2)$ iid.
-The receiver then estimates $\widehat\Theta_i = \widehat\Theta_{i-1} + \frac{Y_i}{i}$, which results in:
+Therefore, in general, the transmitter sends
+\begin{equation}
+	X_i = \Theta - \widehat\Theta_{i-1} \label{eq:tx-x-noiseless}
+\end{equation}
+and the receiver receives
+\begin{equation}
+	Y_i = X_i + N_i
+\end{equation}
+where $N_i \sim \mathcal{N}(0, \sigma_N^2)$ iid. The receiver then estimates
+\begin{equation}
+	\widehat\Theta_i = \widehat\Theta_{i-1} + \frac{Y_i}{i} \label{eq:rx-thetahat-noiseless}
+\end{equation}
+which results in:
 \begin{align*}
@@
 That is,
-\begin{align*}
-	\text{Transmitter's transmissions:}\;X_i &= \Theta - (\widehat\Theta_{i-1} + R_{i-1}) \\
-	\text{Channel outputs at the receiver:}\;	Y_i &= X_i + N_i \\
-	\text{Receiver's estimates \& transmissions:}\;	\widehat\Theta_i &= \widehat\Theta_{i-1} + \frac{Y_i}{i}
-\end{align*}
+\begin{align}
+	& \text{Transmitter's transmissions:}\;X_i = \Theta - (\widehat\Theta_{i-1} + R_{i-1}) \\ \label{eq:noisy-tx-x}
+	& \text{Channel outputs at the receiver:}\;	Y_i = X_i + N_i \\
+	& \text{Receiver's estimates \& transmissions:}\;	\widehat\Theta_i = \widehat\Theta_{i-1} + \frac{Y_i}{i}
+\end{align}
 where $R_{i-1}$ is the AWGN noise in the reverse link. $R_i \sim \mathcal{N}(0, \sigma_R^2)$ are iid random variables.
@@
 
-In order to compute the Granger causality in the reverse direction (from receiver to transmitter), we wish first to compute coefficients $\alpha_j$ such that:
+In order to compute the Granger causality in the reverse direction (from the receiver to the transmitter), we model $X_i$ as a linear function of its past.
 \begin{equation*}
@@
 \end{equation*}
+We then compute coefficients $\alpha_j$ such that the average error in fitting $X_i$ for all $i$ is minimized.
 Note that $\alpha_j$ can themselves depend on $i$, since this is a non-stationary process (because the error $\Theta-\widehat{\Theta}_i = X_i$ converges to zero).
+We describe how these coefficients might be estimated in a more general setting, and justify using theoretically determined system parameters as regression coefficients in section~\ref{sec:estimating-ar-coeffs}.
+
+For now, we theoretically evaluate the system parameters.
+We start with equation~\ref{eq:tx-x-noiseless} and manipulate terms to arrive at an equation bearing the required form of equation~\ref{eq:ar-model-x-x}:
 \begin{align*}
@@
 
-Next, we wish to write $X_i$ in terms of $X_{i-j}$ and $\widehat\Theta_{i-j}$:
-\begin{equation*}
-	X_i = \sum_{j=1}^{p}{\alpha_j X_{i-j}} + \sum_{j=1}^{p}{\beta_j \widehat\Theta_{i-j}} + \tilde\epsilon_i \numberthis \label{eq:ar-model-x-x-theta}
-\end{equation*}
-which gives
+Next, we model $X_i$ in terms of both the past of $X$ and the past of $\widehat\Theta$:
+\begin{equation}
+	X_i = \sum_{j=1}^{p}{\alpha_j X_{i-j}} + \sum_{j=1}^{p}{\beta_j \widehat\Theta_{i-j}} + \tilde\epsilon_i \label{eq:ar-model-x-x-theta}
+\end{equation}
+We can manipulate equation~\ref{eq:tx-x-noiseless} to bring it into the above form:
 \begin{align*}
@@
 
-In the forward direction, we do not explicitly compute the Granger causality ratio, but simply show that it is bounded strictly between 0 and $\infty$:
-\begin{equation*}
-	\widehat\Theta_i = \sum_{j=1}^p{\alpha_j \widehat\Theta_{i-j}} + \epsilon_i
-\end{equation*}
+In the forward direction, we do not explicitly compute the Granger causality ratio, but simply show that it is bounded strictly between 1 and $\infty$:
+\begin{equation}
+	\widehat\Theta_i = \sum_{j=1}^p{\alpha_j \widehat\Theta_{i-j}} + \epsilon_i \label{eq:ar-model-theta-theta}
+\end{equation}
 \begin{align*}
@@
 Further, if we try to predict $\widehat\Theta_i$ from the previous $\widehat\Theta_{i-i}$ and $X_{i-j}$:
-\begin{equation*}
-	\widehat\Theta_i = \sum_{j=1}^{p}{\alpha_j \widehat\Theta_{i-j}} + \sum_{j=0}^{p-1}{\beta_j X_{i-j}} + \tilde\epsilon_i
-\end{equation*}
+\begin{equation}
+	\widehat\Theta_i = \sum_{j=1}^{p}{\alpha_j \widehat\Theta_{i-j}} + \sum_{j=0}^{p-1}{\beta_j X_{i-j}} + \tilde\epsilon_i \label{eq:ar-model-theta-theta-x}
+\end{equation}
 we see that
@@
 
-The intuitive argument for why this is happening might go as follows: since the feedback link is noiseless, you can always perfectly predict the transmitted symbol from the $\widehat\Theta$'s and the history of $X$.
-On the other hand, you can never perfectly predict $\widehat\Theta_i$ from the $X$'s and the history of $\widehat\Theta$.
+The intuitive argument for why this is happening might go as follows: since the feedback link is noiseless, you can always perfectly predict the transmitted symbol from the present and past $\widehat\Theta$'s and the history of $X$.
+On the other hand, you can never perfectly predict $\widehat\Theta_i$ from the present and past $X$'s and the history of $\widehat\Theta$.
 

### 2015-07-07 paper.tex
Fix Pearl reference page numbers.
Add note about linearity.

@@
 \end{equation*}
-We then compute coefficients $\alpha_j$ such that the average error in fitting $X_i$ for all $i$ is minimized.
+We then compute coefficients $\alpha_j$ such that the average error in fitting $X_i$ is minimized.
 Note that $\alpha_j$ can themselves depend on $i$, since this is a non-stationary process (because the error $\Theta-\widehat{\Theta}_i = X_i$ converges to zero).
@@
 With a sufficiently large number of trials, therefore, the system parameters (to be used as regression coefficients in the Granger Causality analysis) can be estimated to arbitrarily high accuracy.
+
+It should be noted that we have restricted ourselves to an analysis of linear strategies: the channel, the extended Schalkwijk and Kailath scheme to a noisy feedback link, and the proposed regression model are all linear.
 

### 2015-07-07 paper.tex
Add GC comparison figure.

@@
 
-In order to determine the direction of causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is \emph{compared} to that from $\{Y\}$ to $\{X\}$.
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+	\centering
+	\includegraphics[width=3.2in]{gc-comparison}
+	\caption{In order to determine the direction of greater causal influence, the Granger Causality metrics in the forward and reverse directions are compared}
+	\label{fig:sk-block-diagram}
+\end{figure}
+
+In order to determine the direction of greater causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is \emph{compared} to that from $\{Y\}$ to $\{X\}$.
 The direction of causal influence is taken to be the direction with a greater Granger Causality metric.

### 2015-07-07 paper.tex
Add figure for GC vs message.

@@
 
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+	\centering
+	\includegraphics[width=3.2in]{gc-vs-message}
+	\caption{Is the direction of stronger Granger-causal influence necessarily the same as the direction in which the information (or equivalently, the ``message'') is flowing?}
+	\label{fig:gc-vs-message}
+\end{figure}
+
 % Intro: How GC is used in neuroscience today ---------------------------------
@@
 	\caption{In order to determine the direction of greater causal influence, the Granger Causality metrics in the forward and reverse directions are compared}
-	\label{fig:sk-block-diagram}
+	\label{fig:gc-comparison}
 \end{figure}
@@
 which means that $0 < \frac{\sigma_N^2}{i^2} < \text{Var}(\epsilon_i) < \frac{\sigma_N^2 + \sigma_\theta^2}{i^2} < \infty$.
-Further, if we try to predict $\widehat\Theta_i$ from the previous $\widehat\Theta_{i-i}$ and $X_{i-j}$:
+Further, if we try to predict $\widehat\Theta_i$ from the previous $\widehat\Theta_{i-j}$ and $X_{i-j}$:
 \begin{equation}
@@
 
-The intuitive argument for why this is happening might go as follows: since the feedback link is noiseless, you can always perfectly predict the transmitted symbol from the present and past $\widehat\Theta$'s and the history of $X$.
-On the other hand, you can never perfectly predict $\widehat\Theta_i$ from the present and past $X$'s and the history of $\widehat\Theta$.
+The intuitive argument for why this is happening might go as follows: since the feedback link is noiseless, you can always perfectly predict the transmitted symbol from the past $\widehat\Theta$'s and the history of $X$.
+On the other hand, you can never perfectly predict $\widehat\Theta_i$ from the past $X$'s and the history of $\widehat\Theta$.
 

### 2015-07-07 paper.tex
Make minor fixes.

@@
 The Granger causality metric is the ratio of these errors, encapsulating the innovations that the process $\{X\}$ causally supplies to the process $\{Y\}$.
-Many variations of Granger causality have also been considered, including a generalization -- Directed Information (see~\cite{MasseyCausality,quinn,weissman}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
-These form possible alternatives for estimating the direction of causal influence, but Directed Information is a generalization of many of these metrics~\cite{quinn}.
+Many variants of Granger causality have also been considered, including a generalization -- Directed Information (see~\cite{MasseyCausality,quinn,weissman}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
+These variants form possible alternatives for estimating the direction of causal influence, but Directed Information is a generalization of many of these metrics~\cite{quinn}.
 
@@
 \end{align}
-where $R_{i-1}$ is the AWGN noise in the reverse link. $R_i \sim \mathcal{N}(0, \sigma_R^2)$ are iid random variables.
+where $R_{i-1}$ is the AWGN noise in the reverse link. $R_i \sim \mathcal{N}(0, \sigma_R^2)$ are iid.\ random variables.
 

### 2015-07-10 paper.tex
Add more acknowledgements.
Make Pulkit's edits.

@@
 \begin{abstract}
-	Granger causality is an established measure of the ``causal influence'' that one stochastic process has on another.
-Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer causal influences.
+	Granger causality is an established statistical measure of the ``causal influence'' that one stochastic process has on another.
+Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer statistical causal influences.
 Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow''.
@@
 We show that in these experiments, the direction of information flow can be opposite to that which is inferred using Granger causality and Directed Information.
-Thus, while it might be reasonable to infer the direction of causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
+Thus, while it might be reasonable to infer the direction of statistical causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
 \end{abstract}
@@
 In order to understand how the brain performs computations, it could be useful to first understand the directions of \emph{information flow} in various parts of the brain (e.g.~\cite{blinowska2004granger,dhamala2008analyzing,nolte2008robustly,korzeniewska2003determination,schippers2010mapping} etc.).
-In an effort to make headway on the goals of \emph{BRAIN 2025}, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how this information flows (eg.~\cite{Brovelli2004BetaOscillations,goebel2003investigating,deshpande2008effective}), or to acquire directed maps of functional connectivity (eg.~\cite{friston2013analysing,goebel2003investigating,deshpande2008effective}).
+In an effort to make headway on the goals of \emph{BRAIN 2025}, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how this information flows (e.g.~\cite{Brovelli2004BetaOscillations,goebel2003investigating,deshpande2008effective}), or to acquire directed maps of functional connectivity (e.g.~\cite{friston2013analysing,goebel2003investigating,deshpande2008effective}).
 For instance, in~\cite{Brovelli2004BetaOscillations}, Granger causal influences that are measured between somatosensory and motor sites are said to ``support the idea that somatosensory feedback provides information to the sensorimotor system that is used to control motor output''.
@@
 
-In order to determine the direction of greater causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is \emph{compared} to that from $\{Y\}$ to $\{X\}$.
-The direction of causal influence is taken to be the direction with a greater Granger Causality metric.
-Further, this direction of causal influence is interpreted to be the direction of information flow -- the interpretation we question in this paper.
+In order to determine the direction of \emph{greater} causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is often \emph{compared} to that from $\{Y\}$ to $\{X\}$.
+The direction of causal influence is then usually taken to be the direction with the greater Granger Causality metric (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal}).
+Further, this direction of causal influence is interpreted to be the direction of information flow, which is the interpretation we question in this paper.
 We also note that many of these works use a spectral version of Granger Causality, that supplies this metric as a function of frequency.
@@
 
-We note here that Granger's original analysis does not compare this metric on forward and reverse links, however, this is commonly done in practice even in areas beyond neuroscience (eg.~\cite{Brovelli2004BetaOscillations,jiao2013universal}).
+We note here that Granger's original analysis does not compare this metric on forward and reverse links, and even the stronger notion of true causality (see remark~\ref{remark:gc-vs-true} in section~\ref{sec:objections-and-shortcomings}) does not involve this comparison.
+However, this is commonly done in practice even in areas beyond neuroscience (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal}).
 
-While we accept that Granger Causal influence can be accurately computed (provided measurements are taken suitably; see Section~\ref{sec:previous-criticisms-of-gc}) and can be useful in many situations (see Section~\ref{sec:conclusions}), interpreting the direction of Granger Causal influence as the direction of information flow can be erroneous, as we demonstrate in this paper.
+While we accept that it might be possible to accurately estimate Granger Causal influence (provided measurements are taken suitably; see Section~\ref{sec:previous-criticisms-of-gc}) and that it could be useful in many situations (see Section~\ref{sec:conclusions}), interpreting the direction of greater Granger Causal influence as the direction of information flow can be erroneous, as we demonstrate in this paper.
 
@@
 
-Several objections have been leveled against Granger Causality in the past.
+Several objections have been raised against the use of Granger Causality in the past.
 We give, here, a short overview of these and describe why our objection is novel, and possibly more fundamental in nature, at least in the context of its usage in neuroscience.
@@
 We believe that this argument is much more serious than previous objections, at least in the context of determining the directions of information flow in neuroscientific experiments, towards understanding the computational functions of brain regions.
+
+A more serious objection has to do with the difference between statistical measures of causality (such as Granger Causality) and true causality, and whether or not our work simply alludes to this difference.
+Our principal argument is different, however, as we describe in remark~\ref{remark:gc-vs-true} in Section~\ref{sec:objections-and-shortcomings}.
 
@@
 \begin{enumerate}
-	\item Firstly, and most importantly, previous work has already shown that presence of a Granger Causal influence does not imply true causation.
-Judea Pearl classifies Granger Causality as stemming from a ``statistical'' model, rather than from a ``causal'' model~\cite{pearl2009causality-gc-stat}.
-A potential issue, therefore, might be the question the novelty of this work, given that Pearl's classification could be seen to clear up the misconception.
-While our argument is, in spirit, very similar to this criticism of Granger causality, in the situation we study, we ask a different question - we wish to understand whether or not Granger Causality gives the correct direction of \emph{information flow}.
-Further, we give a concrete counter-example in this context, which is very relevant in the modern neuroscience setting.
+	\item \label{remark:gc-vs-true} Previous work has already noted that Granger Causal influence does not imply true causation~\cite{granger1980testing}.
+This differentiation is made rigorously by Judea Pearl~\cite{pearl2009causality-gc-stat}, where he classifies Granger Causality as stemming from a ``statistical'' model, rather than from a ``causal'' model.
+The argument we make is very similar in spirit: we ask whether or not Granger Causality gives the correct direction of \emph{information flow}.
+A question on the novelty of our work may therefore be raised: if our argument boils down to a restatement of Pearl's differentiation, then this work has no new conceptual contribution.
+\\
+		We make the case that our argument \emph{is} novel in the following manner: in the systems we consider -- the brain, as well as the communication system in our counter-example -- causal influences exist in both directions.
+In our counter-example, for instance, the feedback communication algorithm that is employed involves transmissions from both $Tx$ and $Rx$.
+The transmissions of each depend on what was transmitted by the other in the previous time instant.
+Causal influence and true causation, therefore, exist in both directions.
+The information of interest, however, flows in only one direction: from the transmitter to the receiver.
+We ask whether or not the direction of \emph{this} information flow can be discerned by comparing Granger causal influences in each direction.
+To this end, we give a concrete counter-example.
+This work is particularly relevant in the context of modern neuroscience, where such directions of information flow are desired in order to understand brain function.
 	\item Our analysis does not tackle an information source that evolves with time.
@@
 This is a clear direction for future work.
-	\item The power and energy constraints are somewhat oversimplified to make the analysis simpler. This is for simplicity of exposition. A more general analysis is a trivial extension.
+	\item The power and energy constraints are somewhat oversimplified to make the analysis simpler. This is for simplicity of exposition. A more general analysis is a simple extension.
 	\item We have also restricted ourselves to analyzing linear feedback communication strategies.
 In the presence of noise in the feedback link, linear communication strategies are known to be sub-optimal~\cite{YoungHanKimPaper}.
-In order to make a water-tight argument, we would need to show that Granger Causality fails to correctly predict the direction of information flow, even when an optimal communication strategy -- which would have to be non-linear.
+In order to make a water-tight argument, we would need to show that Granger Causality fails to correctly predict the direction of information flow, even when an optimal (non-linear) communication strategy is employed.
 This could be scope for future work.
-However, we do not expect results to change dramatically -- when the feedback link is impaired by noise of only very low variance, the noiseless case should make for a good approximation of the system, and results should degrade gracefully, if there is any degradation at all.
+However, we do not expect results to change dramatically: when the feedback link is impaired by noise of only very low variance, the noiseless case should make for a good approximation of the system, and results should degrade gracefully, if there is any degradation at all.
 	\item We consider a simple point-to-point network.
@@
 The authors would like to thank Nicola Elia for discussions that suggested this possible misunderstanding in use of Granger causality and Directed Information in computational systems.
+We also thank Rob Kass, Momin Malik and Kun Zhang for useful discussions.
+Finally, we thank the reviewers of a previous submission of this work to ISIT for their insightful remarks and comments.
 Praveen Venkatesh is partially supported by a dean's fellowship.

### 2015-07-10 paper.tex
Remove short-time-window para from intro.

@@
 
-The computation of Granger causality, in general, requires that the processes under consideration be stationary.
+%The computation of Granger causality, in general, requires that the processes under consideration be stationary.
 In neuroscientific experiments, this is usually not the case, as the express objective of the experiment is to understand brain function while the brain is subjected to a task.

### 2015-07-10 paper.tex
Shorten matrix eqn to bring back to 10 pages.
Fix inter-image spacings.
Make \ref -> \eqref as required.

@@
 For now, we theoretically evaluate the system parameters.
-We start with equation~\ref{eq:tx-x-noiseless} and manipulate terms to arrive at an equation bearing the required form of equation~\ref{eq:ar-model-x-x}:
+We start with equation~\eqref{eq:tx-x-noiseless} and manipulate terms to arrive at an equation bearing the required form of equation~\eqref{eq:ar-model-x-x}:
 \begin{align*}
@@
 \end{equation}
-We can manipulate equation~\ref{eq:tx-x-noiseless} to bring it into the above form:
+We can manipulate equation~\eqref{eq:tx-x-noiseless} to bring it into the above form:
 \begin{align*}
@@
 
-The Granger Causality and Directed Information metrics are a function of the regression coefficients estimated from the data by fitting the models given by equations~\ref{eq:ar-model-x-x} and~\ref{eq:ar-model-x-x-theta}.
+The Granger Causality and Directed Information metrics are a function of the regression coefficients estimated from the data by fitting the models given by equations~\eqref{eq:ar-model-x-x} and~\eqref{eq:ar-model-x-x-theta}.
 In our analysis, we described the data-generation model: the algorithm inspired by the Schalkwijk and Kailath scheme for feedback communication.
@@
 Suppose we record two non-stationary processes, $\{X_t\}_{t=1}^n$ and $\{Y_t\}_{t=1}^n$, for which we seek to compute the Granger Causality metrics.
-To this end, we must find coefficients $\alpha_j(t)$ and $\beta_j(t)$ to minimize the error in fitting the models given by equations~\ref{eq:ar-model-x-x} and~\ref{eq:ar-model-x-x-theta}.
+To this end, we must find coefficients $\alpha_j(t)$ and $\beta_j(t)$ to minimize the error in fitting the models given by equations~\eqref{eq:ar-model-x-x} and~\eqref{eq:ar-model-x-x-theta}.
 Note that $\alpha$ and $\beta$ depend on $t$, since we assume the process is non-stationary.
@@
 	&=
-	\left[ \begin{array}{cccc}
-			X_{t-1}^{(1)} & X_{t-2}^{(1)} & \cdots & X_{t-p}^{(1)} \\
-			X_{t-1}^{(2)} & X_{t-2}^{(2)} & \cdots & X_{t-p}^{(2)} \\
-			\vdots &  & \ddots & \vdots \\
-			X_{t-1}^{(N)} & X_{t-2}^{(N)} & \cdots & X_{t-p}^{(N)} \\
+	\left[ \begin{array}{ccc}
+			X_{t-1}^{(1)} & \cdots & X_{t-p}^{(1)} \\
+			X_{t-1}^{(2)} & \cdots & X_{t-p}^{(2)} \\
+			\vdots & \ddots & \vdots \\
+			X_{t-1}^{(N)} & \cdots & X_{t-p}^{(N)} \\
+			%X_{t-1}^{(1)} & X_{t-2}^{(1)} & \cdots & X_{t-p}^{(1)} \\
+			%X_{t-1}^{(2)} & X_{t-2}^{(2)} & \cdots & X_{t-p}^{(2)} \\
+			%\vdots &  & \ddots & \vdots \\
+			%X_{t-1}^{(N)} & X_{t-2}^{(N)} & \cdots & X_{t-p}^{(N)} \\
 	\end{array} \right]
@@
 			\alpha_p(t)
-	\end{array} \right] \\
-	&\qquad \qquad \qquad +
+	\end{array} \right] +
+	%&\qquad \qquad \qquad +
 	\left[ \begin{array}{c}
@@
 This is an unbiased and consistent estimator for $\bf{\alpha}(t)$.
-This analysis can be trivially extended to the model described by equation~\ref{eq:ar-model-x-x-theta}.
+This analysis can be trivially extended to the model described by equation~\eqref{eq:ar-model-x-x-theta}.
 With a sufficiently large number of trials, therefore, the system parameters (to be used as regression coefficients in the Granger Causality analysis) can be estimated to arbitrarily high accuracy.
@@
 	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point20} \\
+	\vspace{5mm}
 	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point35} \\
+	\vspace{5mm}
 	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point60}

### 2015-07-10 paper.tex
Make miscellaneous edits.

@@
 The Granger causality metric is the ratio of these errors, encapsulating the innovations that the process $\{X\}$ causally supplies to the process $\{Y\}$.
-Many variants of Granger causality have also been considered, including a generalization -- Directed Information (see~\cite{MasseyCausality,quinn,weissman}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
+Many variants of Granger causality have also been developed, including a generalization -- Directed Information (see~\cite{MasseyCausality,quinn,weissman}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
 These variants form possible alternatives for estimating the direction of causal influence, but Directed Information is a generalization of many of these metrics~\cite{quinn}.
@@
 In order to determine the direction of \emph{greater} causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is often \emph{compared} to that from $\{Y\}$ to $\{X\}$.
-The direction of causal influence is then usually taken to be the direction with the greater Granger Causality metric (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal}).
+The direction of causal influence is then taken to be the direction with the greater Granger Causality metric (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal}).
 Further, this direction of causal influence is interpreted to be the direction of information flow, which is the interpretation we question in this paper.
@@
 It then becomes possible to also determine the brain wave frequency at which these influences occur.
-However, we restrict our analysis to the simpler non-spectral version of Granger Causality, since it is sufficient to for the purpose of our arguments.
+However, we restrict our analysis to the simpler non-spectral version of Granger Causality, since it is sufficient for the purpose of our arguments.
 
@@
 It is important to note that the technical objections listed above are all deficiencies in or limitations of \emph{measurement}.
-They indicate that incorrect Granger Causal influences may be computed if there is some deficiency or limitation in the measurement procedure.
+They indicate that incorrect Granger Causal influences may be estimated if there is some deficiency or limitation in the measurement procedure.
 These objections can be resolved by taking better measurements (by sampling more nodes, with higher signal-to-noise ratio, etc.).
@@
 The second experiment allows for noise in the feedback channel.
-In both cases, linear strategies inspired from the scheme of Schalkwijk and Kailath~\cite{S&K} are used.
+In both cases, linear strategies inspired by the scheme of Schalkwijk and Kailath~\cite{S&K} are used.
 
-Our goal here is to bring out the point that whether Granger causality and Directed Information can be used to interpret direction of information flows is an issue that can be, and perhaps should be, considered using thought experiments on simple communication problems where \emph{information flow direction, and quantity, is already known}.
+Our goal here is to bring out the point that whether Granger causality and Directed Information can be used to interpret the direction of information flow is an issue that can be, and perhaps should be, considered using thought experiments on simple communication problems where \emph{information flow direction, and quantity, is already known}.
 If the direction of causal influence yielded by Granger causality or some other similar measure were to match the known direction of information flow, then that measure can be more confidently used in experiments.
@@
 
-A review of a submission to a previous conference raised some objections to this work, which we discuss to clarify our perspective.
+A review of a previous conference submission of this paper had raised some objections to this work, which we discuss here to clarify our perspective.
 Further, our analysis has certain shortcomings, which we acknowledge.
@@
 	\item \label{remark:gc-vs-true} Previous work has already noted that Granger Causal influence does not imply true causation~\cite{granger1980testing}.
-This differentiation is made rigorously by Judea Pearl~\cite{pearl2009causality-gc-stat}, where he classifies Granger Causality as stemming from a ``statistical'' model, rather than from a ``causal'' model.
+This distinction is made rigorously by Judea Pearl~\cite{pearl2009causality-gc-stat}, where he classifies Granger Causality as stemming from a ``statistical'' model, rather than from a ``causal'' model.
 The argument we make is very similar in spirit: we ask whether or not Granger Causality gives the correct direction of \emph{information flow}.
-A question on the novelty of our work may therefore be raised: if our argument boils down to a restatement of Pearl's differentiation, then this work has no new conceptual contribution.
+A question on the novelty of our work may therefore be raised: if our argument boils down to a restatement of Pearl's distinction, then this work has no new conceptual contribution.
 \\
@@
 Hence, our communication process (inspired from the scheme of Schalkwijk and Kailath~\cite{S&K}) is non-ergodic.
-Since Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process, we compute the obvious generalization of Granger causality to non-ergodic processes.
+Since Granger causality is really just relative errors in prediction of a process, in the presence or absence of knowledge of another process, we compute the obvious generalization of Granger causality to non-ergodic processes.
 Nevertheless, future work will address a situation with a linear dynamical system as the information source.
@@
 We do not seek to understate the importance of determining causal influences in the brain; understanding causal influence itself may have a great deal of benefit.
-For instance, we might seek to understand the spread of activity in the brain during an epileptic siezure -- in such applications, we are not concerned with how information is being transferred through the neural circuitry; we only seek to determine the source of the activity for the purpose of surgical intervention.
+For instance, we might seek to understand the spread of activity in the brain during an epileptic seizure -- in such applications, we are not concerned with how information is being transferred through the neural circuitry; we only seek to determine the source of the activity for the purpose of surgical intervention.
 

### 2015-07-10 paper.tex
Edit equation to fit within margins.
Make extreme edits to fit equations in margins.

@@
 The Granger causality metric is the ratio of these errors, encapsulating the innovations that the process $\{X\}$ causally supplies to the process $\{Y\}$.
-Many variants of Granger causality have also been developed, including a generalization -- Directed Information (see~\cite{MasseyCausality,quinn,weissman}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
+Many variants of Granger causality have also been developed, including a generalization -- Directed Information (see~\cite{MasseyCausality,quinn,jiao2013universal}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
 These variants form possible alternatives for estimating the direction of causal influence, but Directed Information is a generalization of many of these metrics~\cite{quinn}.
@@
 We note here that Granger's original analysis does not compare this metric on forward and reverse links, and even the stronger notion of true causality (see remark~\ref{remark:gc-vs-true} in section~\ref{sec:objections-and-shortcomings}) does not involve this comparison.
-However, this is commonly done in practice even in areas beyond neuroscience (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal}).
+However, this is commonly done in practice even in areas beyond neuroscience (e.g.~\cite{Brovelli2004BetaOscillations,hesse2003use,jiao2013universal}).
 
@@
 	\mathbb{E}[(\widehat{\Theta}_{i}-\Theta)^{2}] &= \mathbb{E}\bigg[\bigg(\frac{1}{i}\sum_{j=1}^{i}N_{j}\bigg)^{2}\bigg] \\
-												  &= \frac{1}{i^{2}}\mathbb{E}\bigg[\bigg(\sum_{j=1}^{i}N_{j}\bigg)^{2}\bigg] = \frac{i}{i^{2}}\mathbb{E}[N_{1}^{2}] = \frac{\sigma_{N}^{2}}{i}\overset{i\to\infty}\longrightarrow 0.
+												  &= \frac{1}{i^{2}}\mathbb{E}\bigg[\bigg(\sum_{j=1}^{i}N_{j}\bigg)^{2}\bigg] \\
+												  &= \frac{i}{i^{2}}\mathbb{E}[N_{1}^{2}] = \frac{\sigma_{N}^{2}}{i}\overset{i\to\infty}\longrightarrow 0.
 \end{align*}
@@
 							 &= \frac{1}{2} \log \bigg( 2 \pi e \frac{\sigma_N^2 + \sigma_R^2}{\sigma_R^2} \bigg) \\
-							 & \; \; + \sum_{i=2}^{n-1} \bigg( \frac{1}{2} \log \bigg( 2 \pi e \text{Var} \bigg[ R_{i-1} - \frac{N_i}{i} - R_i \bigg| X^i \bigg] \bigg) \\
+							 & + \sum_{i=2}^{n-1} \bigg( \frac{1}{2} \log \bigg( 2 \pi e \text{Var} \bigg[ R_{i-1} - \frac{N_i}{i} - R_i \bigg| X^i \bigg] \bigg) \\
 							 & \qquad \qquad - \frac{1}{2} \log( 2 \pi e \sigma_R^2 ) \bigg)
@@
 \begin{align*}
-	\left[ \begin{array}{c}
+	\left[ \!\! \begin{array}{c}
 			X_t^{(1)} \\
@@
 			X_t^{(N)}
-	\end{array} \right]
+	\end{array} \!\! \right]
 	&=
-	\left[ \begin{array}{ccc}
+	\left[ \! \begin{array}{ccc}
 			X_{t-1}^{(1)} & \cdots & X_{t-p}^{(1)} \\
@@
 			%X_{t-1}^{(N)} & X_{t-2}^{(N)} & \cdots & X_{t-p}^{(N)} \\
-	\end{array} \right]
-	\left[ \begin{array}{c}
+	\end{array} \! \right]
+	\left[ \!\! \begin{array}{c}
 			\alpha_1(t) \\
@@
 			\alpha_p(t)
-	\end{array} \right] +
+	\end{array} \!\! \right] +
 	%&\qquad \qquad \qquad +
-	\left[ \begin{array}{c}
+	\left[ \!\! \begin{array}{c}
 			\epsilon_t^{(1)} \\
@@
 			\epsilon_t^{(N)}
-	\end{array} \right]
+	\end{array} \!\! \right]
 \end{align*}
@@
 	\mathbb{E}[X_p X_q] &= \mathbb{E} \bigg[ \bigg( \frac{1}{p-1} \sum_{k=1}^{p-2} R_k - \frac{1}{p-1} \sum_{k=1}^{p-1} N_k - R_{p-1} \bigg) \\
-						& \qquad             \bigg( \frac{1}{q-1} \sum_{k=1}^{q-2} R_k - \frac{1}{q-1} \sum_{k=1}^{q-1} N_k - R_{q-1} \bigg) \bigg] \\
-					    &= \frac{\min\{p-2, q-2\}}{(p-1)(q-1)} \sigma_R^2 + \frac{\min\{p-1, q-1\}}{(p-1)(q-1)} \sigma_N^2 \\
+						& \qquad             \bigg( \frac{1}{q-1} \sum_{k=1}^{q-2} R_k - \frac{1}{q-1} \sum_{k=1}^{q-1} N_k - R_{q-1} \bigg) \bigg]
+\end{align*}
+\vspace{-12pt}
+\begin{align*}
+					    &= \frac{\min\{p-2, q-2\}}{(p-1)(q-1)} \sigma_R^2 \\
+						& \qquad + \frac{\min\{p-1, q-1\}}{(p-1)(q-1)} \sigma_N^2 \\
 						& \qquad + \sigma_R^2 \delta_{pq} + \frac{1}{p-1} \sigma_R^2 \mathbb{I}_{p>q} + \frac{1}{q-1} \sigma_R^2 \mathbb{I}_{q>p} \\
 	\mathbb{E}[X_1 X_1] &= \mathbb{E}[\Theta^2] = \sigma_\theta^2 \\
-	\mathbb{E}[X_1 X_j] &= \mathbb{E} \bigg[ \Theta \bigg( \frac{1}{j-1} \sum_{k=1}^{j-2} R_k - \frac{1}{j-1} \sum_{k=1}^{j-1} N_k - R_{j-1} \bigg) \bigg] = 0 \\
+	\mathbb{E}[X_1 X_j] &= \mathbb{E} \bigg[ \Theta \bigg( \frac{1}{j-1} \sum_{k=1}^{j-2} R_k \\
+						& \qquad \qquad \qquad - \frac{1}{j-1} \sum_{k=1}^{j-1} N_k - R_{j-1} \bigg) \bigg] = 0 \\
 	\text{Var}[U|X^i] &= \mathbb{E}[U^2] - \mathbb{E}[U X^i] \mathbb{E}[X^i {X^i}^T]^{-1} \mathbb{E}[X^i U] \\

### 2015-10-04 paper.tex
Add a reference for comparison of GC.

--- 2015-07-10 paper.tex
@@
 \begin{document}
-\title{Is the direction of Granger causal influence the same as the direction of information flow?}
+\title{Is the direction of greater Granger causal influence the same as the direction of information flow?}
 \author{Praveen Venkatesh and Pulkit Grover\\Carnegie Mellon University}
@@
 Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow''.
-We ask: is the direction of causal influence, as predicted by Granger Causality and Directed Information always the same as the direction of information flow?
-We test whether these measures correctly predict the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow (the ``ground truth'') is known by design.
+We ask: is the direction of greater causal influence, as predicted by Granger Causality and Directed Information always the same as the direction of information flow?
+We test whether comparison of these measures in either direction correctly predicts the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow (the ``ground truth'') is known by design.
 The experiments are based on a communication system with a feedback channel, upon which a strategy inspired by the work of Schalkwijk and Kailath is employed.
-We show that in these experiments, the direction of information flow can be opposite to that which is inferred using Granger causality and Directed Information.
+We show that in these experiments, the direction of information flow can be opposite to the direction of greater Granger causal influence or Directed Information.
 Thus, while it might be reasonable to infer the direction of statistical causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
@@
 	\includegraphics[width=3.2in]{gc-comparison}
-	\caption{In order to determine the direction of greater causal influence, the Granger Causality metrics in the forward and reverse directions are compared}
+	\caption{In order to determine the direction of greater causal influence, the Granger Causality metrics in the forward and reverse directions are often compared~\cite{Roebroeck2005MappingDirected}.}
 	\label{fig:gc-comparison}
@@
 In order to determine the direction of \emph{greater} causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is often \emph{compared} to that from $\{Y\}$ to $\{X\}$.
-The direction of causal influence is then taken to be the direction with the greater Granger Causality metric (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal}).
+The direction of causal influence is then taken to be the direction with the greater Granger Causality metric (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal,Roebroeck2005MappingDirected}, see~\cite{Roebroeck2005MappingDirected} for an understanding of what physical constraints motivate this comparison)).
 Further, this direction of causal influence is interpreted to be the direction of information flow, which is the interpretation we question in this paper.
@@
 
-Suppose we try to use the Granger causality measure of directed influence to infer the direction of information flow.
+Suppose we try to use the direction of greater Granger causal influence to infer the direction of information flow.
 As discussed in Section~\ref{sec:intro}, one might expect the direction of information flow -- and hence the direction of greater Granger causality -- to be from the transmitter to the receiver.
@@
 
-In the reverse direction, the Directed Information is infinite.
+In the reverse direction, the Directed Information is $\infty$.
 \begin{align*}

### 2015-10-04 paper.tex
Make multiple minor edits.

@@
 
-In order to understand how the brain performs computations, it could be useful to first understand the directions of \emph{information flow} in various parts of the brain (e.g.~\cite{blinowska2004granger,dhamala2008analyzing,nolte2008robustly,korzeniewska2003determination,schippers2010mapping} etc.).
-In an effort to make headway on the goals of \emph{BRAIN 2025}, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how this information flows (e.g.~\cite{Brovelli2004BetaOscillations,goebel2003investigating,deshpande2008effective}), or to acquire directed maps of functional connectivity (e.g.~\cite{friston2013analysing,goebel2003investigating,deshpande2008effective}).
+In order to understand how the brain performs computations, it could be useful to first understand the directions of \emph{information flow} in various parts of the brain (e.g.~\cite{blinowska2004granger, dhamala2008analyzing, nolte2008robustly, korzeniewska2003determination, schippers2010mapping} etc.).
+In an effort to make headway on the goals of \emph{BRAIN 2025}, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how this information flows (e.g.~\cite{Brovelli2004BetaOscillations, goebel2003investigating, deshpande2008effective}), or to acquire directed maps of functional connectivity (e.g.~\cite{goebel2003investigating, deshpande2008effective, friston2013analysing}).
 For instance, in~\cite{Brovelli2004BetaOscillations}, Granger causal influences that are measured between somatosensory and motor sites are said to ``support the idea that somatosensory feedback provides information to the sensorimotor system that is used to control motor output''.
@@
 
-Several works have outlined the procedures involved in using Granger Causality to estimate causal influences in the brain (\cite{bernasconi2000bidirectional,Brovelli2004BetaOscillations,Ding2000ShortWindow,Roebroeck2005MappingDirected,Bressler2011WienerGranger,Barnett2014MVGC,Ding2006GrangerCausality}).
+Several works have outlined the procedures involved in using Granger Causality to estimate causal influences in the brain (\cite{Brovelli2004BetaOscillations, bernasconi2000bidirectional, Ding2000ShortWindow, Roebroeck2005MappingDirected, Bressler2011WienerGranger, Barnett2014MVGC, Ding2006GrangerCausality}).
 Here, we briefly describe how Granger Causal influence is quantified, and how it is computed in these works.
 
-Granger causality, as originally described by Granger~\cite{Granger1969InvestigatingCausal}, measures the level of influence that one process $\{X\}$ has on another process $\{Y\}$.
-The analysis compares the \emph{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$.
-The Granger causality metric is the ratio of these errors, encapsulating the innovations that the process $\{X\}$ causally supplies to the process $\{Y\}$.
-Many variants of Granger causality have also been developed, including a generalization -- Directed Information (see~\cite{MasseyCausality,quinn,jiao2013universal}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
+Granger causality, as originally described by Granger~\cite{Granger1969InvestigatingCausal}, measures the level of causal influence that one process $\{X\}$ has on another process $\{Y\}$.
+The analysis compares the \emph{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$\footnote{A mathematical exposition of this process appears in Section~\ref{sec:gc-noiseless}}.
+The Granger causality metric is the ratio of these errors, encapsulating the \emph{innovations} that the process $\{X\}$ \emph{causally} supplies to the process $\{Y\}$.
+Many variants of Granger causality have also been developed, including a generalization -- Directed Information (see~\cite{MasseyCausality, quinn, jiao2013universal}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.
 These variants form possible alternatives for estimating the direction of causal influence, but Directed Information is a generalization of many of these metrics~\cite{quinn}.
@@
 In order to determine the direction of \emph{greater} causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is often \emph{compared} to that from $\{Y\}$ to $\{X\}$.
-The direction of causal influence is then taken to be the direction with the greater Granger Causality metric (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal,Roebroeck2005MappingDirected}, see~\cite{Roebroeck2005MappingDirected} for an understanding of what physical constraints motivate this comparison)).
+The direction of causal influence is then taken to be the direction with the greater Granger Causality metric (e.g.~\cite{Brovelli2004BetaOscillations, Roebroeck2005MappingDirected, jiao2013universal}, see~\cite{Roebroeck2005MappingDirected} for an understanding of what physical constraints motivate this comparison).
 Further, this direction of causal influence is interpreted to be the direction of information flow, which is the interpretation we question in this paper.
+We note here that Granger's original analysis does not compare this metric on forward and reverse links, and even the stronger notion of true causality (see remark~\ref{remark:gc-vs-true} in section~\ref{sec:objections-and-shortcomings}) does not involve this comparison.
+However, this is commonly done in practice even in areas beyond neuroscience (e.g.~\cite{Brovelli2004BetaOscillations, Roebroeck2005MappingDirected, jiao2013universal, hesse2003use}).
+
 We also note that many of these works use a spectral version of Granger Causality, that supplies this metric as a function of frequency.
@@
 However, we restrict our analysis to the simpler non-spectral version of Granger Causality, since it is sufficient for the purpose of our arguments.
-
-We note here that Granger's original analysis does not compare this metric on forward and reverse links, and even the stronger notion of true causality (see remark~\ref{remark:gc-vs-true} in section~\ref{sec:objections-and-shortcomings}) does not involve this comparison.
-However, this is commonly done in practice even in areas beyond neuroscience (e.g.~\cite{Brovelli2004BetaOscillations,hesse2003use,jiao2013universal}).
 
@@
 
-Several objections have been raised against the use of Granger Causality in the past.
+Several objections to the use of Granger Causality have been raised in the past.
 We give, here, a short overview of these and describe why our objection is novel, and possibly more fundamental in nature, at least in the context of its usage in neuroscience.
@@
 All nodes need to be observed, therefore, to avoid finding spurious influences.
-	\item Second, if the measurements from each node are differentially affected by noise, then the predicted direction of causal influence might be opposite to the true direction (\cite{nalatore2007mitigating,andersson2005testing}).
+	\item Second, if the measurements from each node are differentially affected by noise, then the predicted direction of causal influence might be opposite to the true direction (\cite{nalatore2007mitigating, andersson2005testing}).
 Measurements need to be relatively noiseless and precise in order to obtain the correct direction of causal influence.
@@
 They indicate that incorrect Granger Causal influences may be estimated if there is some deficiency or limitation in the measurement procedure.
-These objections can be resolved by taking better measurements (by sampling more nodes, with higher signal-to-noise ratio, etc.).
+These objections can be resolved by taking better measurements (by sampling more nodes, using sensors with higher signal-to-noise ratio, etc.).
 
 Our objection, on the other hand, is more fundamental.
-\emph{Even if} the measurements are made with infinite accuracy, and the regression coefficients associated with computing Granger Causality are precisely estimated, and the Granger Causality metric is perfectly computed (as is the case in our counter-examples), Granger Causal influence may \emph{still} not yield the correct direction of information flow.
-To our knowledge, the argument that Granger causal influence can be opposite to the direction of information flow is a novel one.
+\emph{Even if} the measurements are made with infinite accuracy, and the regression coefficients associated with computing Granger Causality are precisely estimated, \emph{and} the Granger Causality metric is perfectly computed (as is the case in our counter-examples), Granger Causal influence may \emph{still} not yield the correct direction of information flow.
+To our knowledge, the argument that greater Granger causal influence can be opposite to the direction of information flow is a novel one.
 We believe that this argument is much more serious than previous objections, at least in the context of determining the directions of information flow in neuroscientific experiments, towards understanding the computational functions of brain regions.
@@
 Provided with these measurements, the experimenter wants to estimate the direction of information flow, which in this context is the direction of flow of the message.
-Our results, derived in Section~\ref{sec:analytical-results} and numerically illustrated in Section~\ref{sec:numerical-results}, show that both Granger causality and Directed Information can incorrectly estimate the direction of information flow.
+Our results, derived in Section~\ref{sec:analytical-results} and numerically illustrated in Section~\ref{sec:numerical-results}, show that the direction of information flow can be incorrectly inferred using both Granger causality and Directed Information.
 The first experiment considers the (unrealistic) case of communication across noiseless feedback channels.
@@
 
- In the first step, the transmitter sends\footnote{We assume that the power constraints are such that the scaling constant $\alpha$ in~\cite{S&K} is $1$.} $X_1 = \Theta$, which the receiver receives with added noise.
+ In the first step, the transmitter sends\footnote{We assume that the power constraints are such that the scaling constant `$\alpha$' in~\cite{S&K} is $1$.} $X_1 = \Theta$, which the receiver receives with added noise.
 The receiver sends back an estimate of $\Theta$ over the feedback link.
@@
 
-Linear strategies are known to be suboptimal for the communication problem~\cite{YoungHanKimPaper} (where $\Theta$ is a quantized random variable communicating a finite-rate message reliably), and for problems with non-classical information structures in general~\cite{Witsenhausen68}.
+Linear strategies are known to be suboptimal for this communication problem~\cite{YoungHanKimPaper} (where $\Theta$ is a quantized random variable communicating a finite-rate message reliably), and for problems with non-classical information structures in general~\cite{Witsenhausen68}.
 Nevertheless, we now show that the resulting estimates $\widehat{\Theta}_i$  still converge to $\Theta$ in mean-square sense:
@@
 \end{align*}
-which means that $0 < \frac{\sigma_N^2}{i^2} < \text{Var}(\epsilon_i) < \frac{\sigma_N^2 + \sigma_\theta^2}{i^2} < \infty$.
+which means that $0 < \frac{\sigma_N^2}{i^2} < \text{Var}(\epsilon_i) < \frac{\sigma_N^2 + \sigma_\theta^2}{i^2} < \infty$, since the past of $\widehat{\Theta}$ cannot be used to explain $N_i$.
 Further, if we try to predict $\widehat\Theta_i$ from the previous $\widehat\Theta_{i-j}$ and $X_{i-j}$:
@@
 
-The intuitive argument for why this is happening might go as follows: since the feedback link is noiseless, you can always perfectly predict the transmitted symbol from the past $\widehat\Theta$'s and the history of $X$.
-On the other hand, you can never perfectly predict $\widehat\Theta_i$ from the past $X$'s and the history of $\widehat\Theta$.
+The intuitive argument for why this is happening might go as follows: since the feedback link is noiseless, one can always perfectly predict the transmitted symbol from the past $\widehat\Theta$'s and the history of $X$.
+On the other hand, one can never perfectly predict $\widehat\Theta_i$ from the past $X$'s and the history of $\widehat\Theta$.
 

### 2015-10-04 paper.tex
Make miscellaneous edits.

@@
 
+\title{Is the direction of greater Granger causal influence the same as the direction of information flow?}
+
+\author{\authorblockN{Praveen Venkatesh\authorrefmark{1} and Pulkit Grover\authorrefmark{2}}
+\authorblockA{Dept.\ of Electrical and Computer Engineering, Carnegie Mellon University \\
+\authorrefmark{1}vpraveen@cmu.edu, \authorrefmark{2}pulkit@cmu.edu}
+}
+
 \begin{document}
-\title{Is the direction of greater Granger causal influence the same as the direction of information flow?}
-\author{Praveen Venkatesh and Pulkit Grover\\Carnegie Mellon University}
 
@@
 \begin{abstract}
-Granger causality is an established statistical measure of the ``causal influence'' that one stochastic process has on another.
+Granger causality is an established statistical measure of the ``causal influence'' that one stochastic process $X$ has on another process $Y$.
 Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer statistical causal influences.
-Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow''.
+More recently, many works compare have begun to compare the Granger causality metrics from $X$ to $Y$ and from $Y$ to $X$, and interpret the direction of causal influence as the ``direction of information flow''.
 We ask: is the direction of greater causal influence, as predicted by Granger Causality and Directed Information always the same as the direction of information flow?
@@
 	\includegraphics[width=3.2in]{gc-vs-message}
-	\caption{Is the direction of stronger Granger-causal influence necessarily the same as the direction in which the information (or equivalently, the ``message'') is flowing?}
+	\caption{Is the direction of stronger Granger-causal influence necessarily the same as the direction in which the message is flowing?}
 	\label{fig:gc-vs-message}
@@
 Causal influence and true causation, therefore, exist in both directions.
-The information of interest, however, flows in only one direction: from the transmitter to the receiver.
+The message, however, flows in only one direction: from the transmitter to the receiver.
 We ask whether or not the direction of \emph{this} information flow can be discerned by comparing Granger causal influences in each direction.
@@
 \end{equation*}
-where $n$ is the number of iterations of the Schalkwijk-Kailath algorithm. For a proof, refer appendix~\ref{app:dir-info-fwd-noiseless}.
+where $n$ is the number of iterations of the Schalkwijk-Kailath algorithm.
+For a proof, refer appendix~\ref{app:dir-info-fwd-noiseless}, which appears in the full version of this paper~\cite{Allerton15Online}.
 
@@
 \end{align*}
-For a derivation of this, see appendix~\ref{app:dir-info-fwd-noisy}.
-
-In the reverse direction,
+For a derivation of this, see appendix~\ref{app:dir-info-fwd-noisy} which appears in~\cite{Allerton15Online}. In the reverse direction,
 \begin{align*}
@@
 \end{align*}
-For a derivation of this, see appendix~\ref{app:dir-info-rev-noisy}.
+For a derivation of this, see appendix~\ref{app:dir-info-rev-noisy} which appears in~\cite{Allerton15Online}.
 

### 2015-10-04 paper.tex
Edit abstract.

@@
 Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer statistical causal influences.
-More recently, many works compare have begun to compare the Granger causality metrics from $X$ to $Y$ and from $Y$ to $X$, and interpret the direction of causal influence as the ``direction of information flow''.
-We ask: is the direction of greater causal influence, as predicted by Granger Causality and Directed Information always the same as the direction of information flow?
-We test whether comparison of these measures in either direction correctly predicts the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow (the ``ground truth'') is known by design.
-The experiments are based on a communication system with a feedback channel, upon which a strategy inspired by the work of Schalkwijk and Kailath is employed.
+More recently, many works compare the Granger causality metrics along forward and reverse links (from $X$ to $Y$ and from $Y$ to $X$), and interpret the direction of greater causal influence as the ``direction of information flow''.
+In this paper, we question whether the direction yielded by comparing Granger Causality or Directed Information along forward and reverse links is always the same as the direction of information flow.
+We explore this question using two simple theoretical experiments, in which the true direction of information flow (the ``ground truth'') is known by design.
+The experiments are based on a communication system with a feedback channel, and employ a strategy inspired by the work of Schalkwijk and Kailath.
 We show that in these experiments, the direction of information flow can be opposite to the direction of greater Granger causal influence or Directed Information.
-Thus, while it might be reasonable to infer the direction of statistical causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
+We also provide information-theoretic intuition for why such counterexamples are not surprising, and why Granger causality-based information-flow inferences will only get more tenuous in larger networks.
+We conclude that one must not use comparison/difference of Granger causality to infer the direction of information flow.
 \end{abstract}

### 2015-10-04 paper.tex
Add acknowledgements.

@@
 
-Suppose we try to use the direction of greater Granger causal influence to infer the direction of information flow.
+%Suppose we try to use the direction of greater Granger causal influence to infer the direction of information flow.
 As discussed in Section~\ref{sec:intro}, one might expect the direction of information flow -- and hence the direction of greater Granger causality -- to be from the transmitter to the receiver.
@@
 \end{align*}
-as expected.
+%as expected.
 
@@
 \end{align*}
-For a derivation of this, see appendix~\ref{app:dir-info-fwd-noisy} which appears in~\cite{Allerton15Online}. In the reverse direction,
+For a derivation of this, see appendix~\ref{app:dir-info-fwd-noisy} which appears in~\cite{Allerton15Online}.
+
+In the reverse direction,
 \begin{align*}
@@
 Note that $\alpha_j(t)$ does not depend on $i$. Collecting variables across $N$ trials, we can write the full model in vector form:
+
 \begin{align*}
@@
 	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point20} \\
-	\vspace{5mm}
+	%\vspace{5mm}
 	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point35} \\
-	\vspace{5mm}
+	%\vspace{5mm}
 	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point60}
@@
 As a final remark, we emphasize that this work only demonstrates the error in \emph{interpreting} the direction of causal influence as the direction of \emph{information flow}.
-We do \emph{not} seek to invalidate much of the neuroscientific work that has been done in this direction; we merely caution against making (what might be construed as hopeful) extrapolations from causal influences as to the manner in which information flows.
+We do \emph{not} seek to invalidate much of the neuroscientific work that has been done in this direction; we merely caution against making (what might be construed as hopeful) extrapolations from causal influences to information flows.
 
@@
 The authors would like to thank Nicola Elia for discussions that suggested this possible misunderstanding in use of Granger causality and Directed Information in computational systems.
-We also thank Rob Kass, Momin Malik and Kun Zhang for useful discussions.
-Finally, we thank the reviewers of a previous submission of this work to ISIT for their insightful remarks and comments.
-Praveen Venkatesh is partially supported by a dean's fellowship.
-%Pulkit Grover is supported by [...].
+We also thank Rob Kass, Bruno Sinopoli, Momin Malik and Kun Zhang for useful discussions.
+Finally, we thank the reviewers of a previous conference submission of this work for their insightful remarks and comments.
+We also thank the CMU BrainHUB initiative for support through a Proseed-BrainHUB seed grant.
+Pulkit Grover was supported through a CMU startup grant, and Praveen Venkatesh through a CMU CIT Dean's fellowship and a CMU Presidential Fellowship.
 

### 2015-10-04 paper.tex
Correct affiliations and references.

@@
 
+\usepackage{cite}
+
 \hyphenation{op-tical net-works semi-conduc-tor}
@@
 \author{\authorblockN{Praveen Venkatesh\authorrefmark{1} and Pulkit Grover\authorrefmark{2}}
-\authorblockA{Dept.\ of Electrical and Computer Engineering, Carnegie Mellon University \\
+\authorblockA{Electrical \& Computer Engineering, and the Center for the Neural Basis of Cognition, Carnegie Mellon University \\
 \authorrefmark{1}vpraveen@cmu.edu, \authorrefmark{2}pulkit@cmu.edu}
@@
 
-The transmitter wants to convey a single zero-mean random number, $\Theta$, having variance
-$\sigma_\theta^2$, to the receiver.
+The transmitter wants to convey a single zero-mean random number, $\Theta$, having variance $\sigma_\theta^2$, to the receiver.
 $\Theta$ could be obtained, for instance, by quantizing a bounded interval on the real line (e.g.
@@
 Therefore, the error in estimating $\Theta$ converges to $0$ in mean-square sense (\textit{i.e.}, $\widehat{\Theta}_i$ is a consistent estimate of $\Theta_i$ even in the presence of noise on the feedback link).
-
-We also derive here an expression for $X_i$, which we will use later.
-\begin{align*}
-	X_i &= \Theta - \widehat\Theta_{i-1} - R_{i-1} \\
-	&= \Theta - \bigg( \Theta + \frac{1}{i-1} \sum_{k=1}^{i-1} N_k - \frac{1}{i-1} \sum_{k=1}^{i-2} R_k \bigg) - R_{i-1} \\
-	&= \frac{1}{i-1} \sum_{k=1}^{i-2} R_k - \frac{1}{i-1} \sum_{k=1}^{i-1} N_k - R_{i} \numberthis \label{eq:x_i-noisy}
-\end{align*}
 
@@
 \end{align*}
-For a derivation of this, see appendix~\ref{app:dir-info-fwd-noisy} which appears in~\cite{Allerton15Online}.
-
-In the reverse direction,
+For a derivation of this, see appendix~\ref{app:dir-info-fwd-noisy} which appears in~\cite{Allerton15Online}. In the reverse direction,
 \begin{align*}
@@
 Note that $\alpha_j(t)$ does not depend on $i$. Collecting variables across $N$ trials, we can write the full model in vector form:
-
 \begin{align*}
@@
 Finally, we thank the reviewers of a previous conference submission of this work for their insightful remarks and comments.
-We also thank the CMU BrainHUB initiative for support through a Proseed-BrainHUB seed grant.
+We thank the CMU BrainHUB initiative for support through a Proseed-BrainHUB seed grant.
 Pulkit Grover was supported through a CMU startup grant, and Praveen Venkatesh through a CMU CIT Dean's fellowship and a CMU Presidential Fellowship.
+
+\IEEEtriggeratref{8}
 
@@
 
+First, we derive an expression for $X_i$, which we will use later.
+\begin{align*}
+	X_i &= \Theta - \widehat\Theta_{i-1} - R_{i-1} \\
+	&= \Theta - \bigg( \Theta + \frac{1}{i-1} \sum_{k=1}^{i-1} N_k - \frac{1}{i-1} \sum_{k=1}^{i-2} R_k \bigg) - R_{i-1} \\
+	&= \frac{1}{i-1} \sum_{k=1}^{i-2} R_k - \frac{1}{i-1} \sum_{k=1}^{i-1} N_k - R_{i} \numberthis \label{eq:x_i-noisy}
+\end{align*}
+
 \begin{align*}
